---
title: Add gene lengths to rowData
author: Gennaro Calendo
format: html
---

The goal of this script will be to create rowData for the gene lengths for all of the 
features included in the database. For TEs it's a little unclear how this should be 
performed at the subfamily level. 

## Set up 

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(GenomicRanges))
suppressPackageStartupMessages(library(GenomicFeatures))

datafiles <- here("analysis", "results", "data-files", "01")
figures <- here("analysis", "results", "figures", "01")
rdsfiles <- here("analysis", "results", "rds-files", "01")

invisible(sapply(c(datafiles, figures, rdsfiles), dir.create, showWarnings=F, recursive=T))
```

## Read in the genome annotation

The REdiscoverTE pipeline uses GENCODE v26 annotations. Download the actual gtf file so that I can create txdbs to compute
exon and transcript lengths 

```{r}
gtffile <- here(datafiles, "gencode.v26.annotation.gtf.gz")

if (!file.exists(gtffile)) {
    download.file(
        "https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_26/gencode.v26.annotation.gtf.gz", 
        gtffile
    )
    gtf <- rtracklayer::import(gtffile)
} else {
    gtf <- rtracklayer::import(gtffile)
}
```

### Total gene length

The position of the start to the end **including introns**

```{r}
genes <- gtf[gtf$type == "gene"]
total_gene_length <- width(genes)
names(total_gene_length) <- genes$gene_id
```

### Sum of reduced exons 

Take all of the unique exon ranges for each gene and sum the total widths

```{r}
txdb <- txdbmaker::makeTxDbFromGRanges(gtf)

exons_by_gene <- exonsBy(txdb, by = "gene")
reduced_exon_lengths <- sum(width(reduce(exons_by_gene)))
names(reduced_exon_lengths) <- names(exons_by_gene)
```

### Median transcript length

Takes the median length of all transcripts for a given gene

```{r}
tx_by_gene <- transcriptsBy(txdb, by = "gene")
median_tx_lengths <- median(width(tx_by_gene))
names(median_tx_lengths) <- names(tx_by_gene)
```

## Create a final table to annotate these genes

```{r}
df <- data.table(
    gene_id = genes$gene_id,
    gene_name = genes$gene_name,
    gene_type = genes$gene_type,
    total_gene_length = total_gene_length[genes$gene_id],
    reduced_exon_lengths = reduced_exon_lengths[genes$gene_id],
    median_tx_lengths = median_tx_lengths[genes$gene_id]
)

fwrite(df, here(datafiles, "v26-gene-length-annotation.tsv"), sep="\t")
```