---
title: What is the relationship between gene length and dysregulation?
author: Gennaro Calendo
format: html
---

## Set up 

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(SummarizedExperiment))

datafiles <- here("analysis", "results", "data-files", "02")
figures <- here("analysis", "results", "figures", "02")
rdsfiles <- here("analysis", "results", "rds-files", "02")

invisible(sapply(c(datafiles, figures, rdsfiles), dir.create, showWarnings=F, recursive=T))
```

## Load the database backend

```{r}
se <- HDF5Array::loadHDF5SummarizedExperiment("/home/gennaro/data/cancer-rnaseq-database/appdata/se_hdf5")

# Subset for only genes 
se <- se[rowData(se)$feature_type == "Gene", ]

# Realize the data into memory since I'm on
se2 <- SummarizedExperiment(
    assays = list(
        "FDR" = as.matrix(assay(se, "adj.P.Val")),
        "logFC" = as.matrix(assay(se, "logFC"))
        ),
    colData = colData(se),
    rowData = rowData(se) 
)
```

## Compute gene length categories

```{r}
q_tol <- Hmisc::cut2(rowData(se2)$total_gene_length, g=4)
q_exl <- Hmisc::cut2(rowData(se2)$reduced_exon_lengths, g=4)
q_mxl <- Hmisc::cut2(rowData(se2)$median_tx_lengths, g=4)
```

## Classify the differential expression results for each experiment

```{r}
up_m <- assay(se2, "logFC") > 0 & assay(se2, "FDR") < 0.05
down_m <- assay(se2, "logFC") < 0 & assay(se2, "FDR") < 0.05

# Mask the missing values in each with NAs
up_m[is.na(as.matrix(assay(se, "AveExpr")))] <- NA
down_m[is.na(as.matrix(assay(se, "AveExpr")))] <- NA

# Also get the number of features measured in each
n_m <- colSums(!is.na(assay(se, "AveExpr")))
```

Now I want to count the values in each column by their gene length category

```{r}
countDeByLength <- function(up, down, l) {
    
    # Compute counts for each length category
    up_by_l <- apply(up, 2, \(x) tapply(x, l, sum, na.rm=TRUE), simplify=FALSE)
    down_by_l <- apply(down, 2, \(x) tapply(x, l, sum, na.rm=TRUE), simplify=FALSE)

    # Bind into data.tables
    up_by_l <- as.data.table(do.call(rbind, up_by_l), keep.rownames="id")
    down_by_l <- as.data.table(do.call(rbind, down_by_l), keep.rownames="id")

    # Add total library sizes and compute percentages
    up_by_l[, lib_size := n_m][, 
                names(.SD) := lapply(.SD, \(x) x / lib_size * 100), .SDcols = names(up_by_l) %like% "\\["][, 
                lib_size := NULL]
    down_by_l[, lib_size := n_m][, 
                names(.SD) := lapply(.SD, \(x) x / lib_size * 100), .SDcols = names(down_by_l) %like% "\\["][, lib_size := NULL]

    # Bind all into a single data.table
    dt <- rbindlist(list("up" = up_by_l, "down" = down_by_l), idcol="direction")

    # Make percentage up/down separate columns with one row for each length category/id
    dt <- melt(dt, id.vars=c("id", "direction"))
    result <- dcast(dt, id + variable ~ direction, value.var = "value")

    # Add the ratio of up:down for each length
    result[, up_down_ratio := up / down]

    return(result)
}
```

Compute the percentages for each of the types of length computations. **These might not be super useful** but hold onto
the calculation for now

```{r}
dt_tol <- countDeByLength(up_m, down_m, q_tol)
dt_exl <- countDeByLength(up_m, down_m, q_exl)
dt_mxl <- countDeByLength(up_m, down_m, q_mxl)
```

## How about probability of DE as a function of gene length?

goseq will fit a probability weight function that computes the probability that a gene is DE based on its length alone

```{r}
# Function used to extract the PWF from the DE + gene lengths
getPwfByLenth <- function(x, l) {
    na_idx <- is.na(x)
    x <- x[!na_idx]
    l <- l[!na_idx]

    result <- goseq::nullp(x, genome = "hg38", id = "geneSymbol", bias.data = l, plot.fit = FALSE)
    result <- result[order(result$bias.data), ]
    result$log10_l <- log10(result$bias.data)

    return(result)
}
```

Get the pwf for each contrast

```{r}
up_pwf_by_tol <- parallel::mclapply(1:ncol(up_m), function(i) {
    getPwfByLenth(up_m[, i], l=rowData(se2)$total_gene_length)
    }, mc.cores=32
)

down_pwf_by_tol <- parallel::mclapply(1:ncol(down_m), function(i) {
    getPwfByLenth(down_m[, i], l=rowData(se2)$total_gene_length)
    }, mc.cores=32
)
```

Show a plot of all contrasts 

```{r}
png(here(figures, "pwf-by-tol.png"), width=12, height=6, units="in", res=300)

par(mfrow=c(1, 2))

tol_rng <- range(log10(rowData(se2)$total_gene_length)) 

# Up-regulation
plot.new()
plot.window(xlim = tol_rng, ylim = c(0,1))
axis(1, lwd = 0, font.axis=2)
axis(2, lwd = 0, font.axis=2)
title(xlab = "Gene Length (log10)")
title(ylab = "Probability Differentially Expressed")
title(main = "Up-regulation", adj = 0)
for (i in 1:length(up_pwf_by_tol)) {
    d <- up_pwf_by_tol[[i]]
    lines(x = d$log10_l, y=d$pwf, col="#00000030")
}

# Down-regulation
plot.new()
plot.window(xlim = tol_rng, ylim = c(0,1))
axis(1, lwd = 0, font.axis=2)
axis(2, lwd = 0, font.axis=2)
title(xlab = "Gene Length (log10)")
title(ylab = "Probability Differentially Expressed")
title(main = "Down-regulation", adj = 0)
for (i in 1:length(down_pwf_by_tol)) {
    d <- down_pwf_by_tol[[i]]
    lines(x = d$log10_l, y=d$pwf, col="#00000030")
}

dev.off()
```


