---
title: "Pool all sample-level counts into a single object"
author: Gennaro Calendo
format: html
---

## Set up

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(patchwork))
suppressPackageStartupMessages(library(paletteer))


datafiles <- here("analysis", "results", "data-files", "03")
figures <- here("analysis", "results", "figures", "03")
rdsfiles <- here("analysis", "results", "rds-files", "03")

invisible(sapply(
  c(datafiles, figures, rdsfiles),
  dir.create,
  showWarnings = FALSE,
  recursive = TRUE
))
```

## Get a list of all processed data

These are the `MultiAssayExperiments` containing tximport counts

```{r}
maefiles <- list.files(
  path = here("bioprojects"),
  pattern = "MultiAssayExperiment.rds",
  full.names = TRUE,
  recursive = TRUE
)
names(maefiles) <- regmatches(maefiles, regexpr("PRJNA[0-9]+", maefiles))
```

## Extract counts and metadata

```{r}
extractData <- function(x) {
  mae <- readRDS(x)
  genes <- MultiAssayExperiment::assay(mae, "gene counts")
  repeats <- MultiAssayExperiment::assay(mae, "RE counts")
  counts <- rbind(genes, repeats)
  metadata <- MultiAssayExperiment::colData(mae)

  return(list(counts = counts, metadata = metadata))
}
```

Extract all counts and metadata into a single sample-level `SummarizedExperiment`

```{r}
run <- FALSE
if (run) {
  suppressPackageStartupMessages(library(MultiAssayExperiment))

  data <- parallel::mclapply(maefiles, extractData, mc.cores = 32)

  all_counts <- lapply(data, `[[`, "counts")
  all_counts <- do.call(cbind, all_counts)

  all_metadata <- lapply(data, `[[`, "metadata")
  all_metadata <- lapply(all_metadata, \(x) {
    as.data.table(data.frame(x), keep.rownames = "Run")
  })
  all_metadata <- rbindlist(all_metadata, fill = TRUE)
  setDF(all_metadata, rownames = all_metadata$Run)

  run_se <- SummarizedExperiment(
    assays = list(
      counts = all_counts,
      logcounts = edgeR::cpm(all_counts, log = TRUE)
    ),
    colData = all_metadata
  )

  saveRDS(run_se, here(rdsfiles, "run-se.rds"))
}
```

## Aggragate to the BioSample level

Sum counts to the biosample level for each run

```{r}
run <- FALSE
if (run) {
  # Read in the run-level data
  run_se <- readRDS(here(rdsfiles, "run-se.rds"))

  # Sum count for each run to the BioSample level
  counts <- sumTechReps(assay(run_se, "counts"), ID = run_se$BioSample)

  # Collapse the metadata to the BioSample level as well
  metadata <- as.data.table(data.frame(colData(run_se)))
  keep_cols <- c(
    "Run",
    "BioSample",
    "BioProject",
    "library_types",
    "frag_dist_length",
    "frag_length_mean",
    "num_eq_classes",
    "num_processed",
    "num_mapped",
    "num_decoy_fragments",
    "num_dovetail_fragments",
    "num_fragments_filtered_vm",
    "num_alignments_below_threshold_for_mapped_fragments_vm",
    "percent_mapped",
    "group"
  )
  metadata <- metadata[, ..keep_cols]
  metadata <- metadata[,
    .(
      library_types = paste(unique(library_types), collapse = ";"),
      avg_frag_dist_length = mean(frag_dist_length),
      avg_frag_length_mean = mean(frag_length_mean),
      avg_num_eq_classes = mean(num_eq_classes),
      avg_num_processed = mean(num_processed),
      avg_num_mapped = mean(num_mapped),
      avg_num_decoy_fragments = mean(num_decoy_fragments),
      avg_num_dovetail_fragments = mean(num_dovetail_fragments),
      avg_num_fragments_filtered_vm = mean(num_fragments_filtered_vm),
      avg_num_alignments_below_threshold_for_mapped_fragments_vm = mean(
        num_alignments_below_threshold_for_mapped_fragments_vm
      ),
      avg_percent_mapped = mean(percent_mapped),
      N_runs = .N
    ),
    by = .(BioSample, BioProject, group)
  ]

  # Classify as single or paired-end libraries
  metadata[,
    library_layout := data.table::fifelse(
      library_types %like% "^[IOM]",
      "PAIRED",
      "SINGLE"
    )
  ]

  setDF(metadata, rownames = metadata$BioSample)
  metadata <- metadata[colnames(counts), ]

  # Save as SummarizedExperiment
  se <- SummarizedExperiment(
    assays = list(counts = counts, logcounts = cpm(counts, log = TRUE)),
    colData = metadata
  )

  saveRDS(se, here(rdsfiles, "biosample-se.rds"))
} else {
  se <- readRDS(here(rdsfiles, "biosample-se.rds"))
}
```

## Show alignment stat distributions

```{r}
getMadCutoff <- function(x, threshold = 3) {
  MAD <- mad(x, na.rm = TRUE)
  cmedian <- median(x, na.rm = TRUE)
  lower <- cmedian - (threshold * MAD)
  upper <- cmedian + (threshold * MAD)
  return(list(upper = upper, lower = lower))
}

plotStat <- function(x, title, paired = TRUE, log_stat = TRUE, ...) {
  if (isTRUE(paired)) {
    X <- colData(se)[se$library_layout == "PAIRED", ][[x]]
  } else {
    X <- colData(se)[se$library_layout == "SINGLE", ][[x]]
  }

  if (log_stat) {
    X <- log10(X)
  }

  hist(
    X,
    breaks = 50,
    main = title,
    col = "cornflowerblue",
    border = "white",
    ylab = "Number of BioSamples",
    ...
  )
  abline(v = getMadCutoff(X)[[1]], col = "red", lty = 2)
  abline(v = getMadCutoff(X)[[2]], col = "red", lty = 2)
}
```

Show histograms of the mapping stats

```{r}
# png(
#   here(figures, "paired-mapping-stats.png"),
#   width = 10,
#   height = 6,
#   units = "in",
#   res = 500
# )
pdf(
  here(figures, "paired-mapping-stats.pdf"),
  width = 10,
  height = 6
)
par(mfrow = c(2, 3))
plotStat(
  "avg_percent_mapped",
  "Percentage Mapped",
  log_stat = FALSE,
  xlab = "Percent Mapped"
)
plotStat("avg_num_mapped", "Number Mapped", xlab = "Number Mapped (log10)")
plotStat(
  "avg_num_eq_classes",
  "Number Equivalence Classes",
  xlab = "Number Eq. Classes (log10)"
)
plotStat(
  "avg_num_decoy_fragments",
  "Number Decoy Fragments",
  xlab = "Number Decoy Fragments (log10)"
)
plotStat(
  "avg_num_dovetail_fragments",
  "Number Dovetail Fragments",
  xlab = "Number Dovetail Fragments (log10)"
)
plotStat(
  "avg_num_fragments_filtered_vm",
  "Number Filtered Fragments",
  xlab = "Number Filtered Fragments (log10)"
)
dev.off()

# png(
#   here(figures, "single-mapping-stats.png"),
#   width = 10,
#   height = 6,
#   units = "in",
#   res = 500
# )
pdf(
  here(figures, "single-mapping-stats.pdf"),
  width = 10,
  height = 6
)
par(mfrow = c(2, 3))
plotStat(
  "avg_percent_mapped",
  "Percentage Mapped",
  paired = FALSE,
  log_stat = FALSE,
  xlab = "Percent Mapped"
)
plotStat(
  "avg_num_mapped",
  "Number Mapped",
  paired = FALSE,
  xlab = "Number Mapped (log10)"
)
plotStat(
  "avg_num_eq_classes",
  "Number Equivalence Classes",
  paired = FALSE,
  xlab = "Number Eq. Classes (log10)"
)
plotStat(
  "avg_num_decoy_fragments",
  "Number Decoy Fragments",
  paired = FALSE,
  xlab = "Number Decoy Fragments (log10)"
)
plotStat(
  "avg_num_fragments_filtered_vm",
  "Number Filtered Fragments",
  paired = FALSE,
  xlab = "Number Filtered Fragments (log10)"
)
dev.off()
```

## Flag and remove outliers based on mapping stats

Outliers should be flagged by library layout type. Convert metadata to data.table for grouping.

```{r}
coldata <- as.data.table(data.frame(colData(se)))
coldata[,
  `:=`(
    avg_num_mapped_outlier = coriell::outliers_by_mad(log10(avg_num_mapped), direction = "low"),
    avg_percent_mapped_outlier = coriell::outliers_by_mad(avg_percent_mapped, direction = "low"),
    avg_num_eq_classes_outlier = coriell::outliers_by_mad(
      log10(avg_num_eq_classes),
      direction = "low"
    ),
    avg_num_decoy_fragments_outlier = coriell::outliers_by_mad(log10(avg_num_decoy_fragments)),
    avg_num_fragments_filtered_vm_outlier = coriell::outliers_by_mad(
      log10(avg_num_fragments_filtered_vm),
      direction = "high"
    )
  ),
  by = library_layout
]

# Flag is number or percentage mapped is too low
coldata[, outlier := (avg_num_mapped_outlier | avg_percent_mapped_outlier)]

# Assign back to SE object
stopifnot(all(coldata$BioSample == colnames(se)))
se <- se[, coldata$outlier == FALSE]
```

## Determine control samples

I will have to use the design and contrasts matrices to determine which samples were used as 
controls. All of the objects have a fit object that contains both a contrasts matrix and a design 
matrix

```{r}
result_sefiles <- list.files(
  path = here("bioprojects"),
  pattern = "se.rds",
  recursive = TRUE,
  full.names = TRUE
)

names(result_sefiles) <- regmatches(
  result_sefiles,
  regexpr("PRJNA[0-9]+", result_sefiles)
)
```

I will need separate functions for designs with intercepts vs designs without.

```{r}
# Test cases with and without intercepts
test_wi <- metadata(readRDS(result_sefiles[[6]]))$fit
test_wo <- metadata(readRDS(result_sefiles[["PRJNA318019"]]))$fit

# Extract the BioSample IDs for controls samples with a no intercept design
getControlsWithoutIntercept <- function(x) {
  cons <- colnames(x$contrasts)
  groups <- rownames(x$contrasts)
  is_control <- apply(x$contrasts, 2, \(x) which(x < 0), simplify = TRUE)

  if (is.list(is_control)) {
    is_control <- unique(unlist(is_control))
  }

  control_groups <- unique(groups[is_control])
  control_design <- x$design[, control_groups, drop = F][
    rowSums(x$design[, control_groups, drop = F]) > 0,
    ,
    drop = F
  ]
  control_samples <- rownames(control_design)

  return(control_samples)
}

# Test case
getControlsWithoutIntercept(test_wo)

# Extract the BioSample IDs for controls with intercept models
getControlsWithIntercept <- function(x) {
  cons <- colnames(x$contrasts)
  groups <- rownames(x$contrasts[rowSums(x$contrasts) > 0, ])
  trt_design <- x$design[, groups, drop = F]
  control_samples <- rownames(trt_design[rowSums(trt_design) == 0, , drop = F])

  return(control_samples)
}

# Test case
getControlsWithIntercept(test_wi)
```

Extract BioSample IDs of control samples.

```{r}
getControls <- function(x) {
  obj <- readRDS(x)
  fit <- metadata(obj)$fit

  tryCatch(
    {
      if ("(Intercept)" %in% colnames(fit$design)) {
        controls <- getControlsWithIntercept(fit)
      } else {
        controls <- getControlsWithoutIntercept(fit)
      }

      return(controls)
    },
    error = function(e) {
      print("Something went wrong. Returning NULL value", e)
      return(NULL)
    }
  )
}

control_file <- here(rdsfiles, "controls.rds")
if (!(file.exists(control_file))) {
  controls <- lapply(result_sefiles, getControls)
  controls <- unlist(controls)
  saveRDS(controls, control_file)
} else {
  controls <- readRDS(control_file)
}
```

### Filter the se object for control samples only

```{r}
se <- se[, se$BioSample %chin% controls]
```

## Map cell types back to BioSamples

All of the contrasts have been annotated with the cell types they were derived from. Some may 
have multiple cell types, i.e. contrasts testing for an overall effect. These can be removed. To 
map back from the contrast metadata (which is curated) to the BioSample level (minimally curated), 
we need to link contrasts to BioSamples. I'll have to create tables mapping contrast -> group 
-> BioSample. The contrast map will contain many more rows than the contrasts in the database 
because all samples within a particular group will be replicated and some samples may belong to 
more than one group.

```{r}
getContrastMapping <- function(x) {
  obj <- readRDS(x)
  fit <- metadata(obj)$fit

  # Get the unique mapping between contrasts and groups
  contrasts <- as.data.table(fit$contrasts, keep.rownames = "group")
  contrasts <- melt(
    contrasts,
    id.vars = "group",
    variable.factor = FALSE,
    variable.name = "contrast"
  )
  contrasts <- unique(contrasts[value != 0, .(group, contrast)])

  # Get the mapping between BioSamples and groups
  design <- as.data.table(fit$design, keep.rownames = "BioSample")
  design <- melt(
    design,
    id.vars = "BioSample",
    variable.factor = FALSE,
    variable.name = "group"
  )
  design <- design[value == 1L, .(BioSample, group)]

  # Map Contrasts -> groups -> BioSamples
  result <- merge(
    contrasts,
    design,
    by = "group",
    all.x = TRUE,
    allow.cartesian = TRUE
  )

  setorder(result, BioSample)
}

contrast_map_file <- here(rdsfiles, "contrast-map.rds")
if (!file.exists(contrast_map_file)) {
  contrast_map <- lapply(result_sefiles, getContrastMapping)
  contrast_map <- rbindlist(contrast_map, idcol = "BioProject")
  contrast_map[, id := paste(BioProject, contrast, sep = ".")]
  saveRDS(contrast_map, contrast_map_file)
} else {
  contrast_map <- readRDS(contrast_map_file)
}
```

## Get the contrast metadata and map information onto BioSamples
 
```{r}
# We only need the metadata shell, not the entire SE object
db <- readRDS("appdata/se_hdf5/se.rds")
contrast_metadata <- as.data.table(data.frame(colData(db)))

# Remove contrasts with multiple cell lines
contrast_metadata <- contrast_metadata[cell_line != "MULTIPLE"]

# Map contrast metadata to BioSample information - some may be dropped that were QC filtered
biosample_metadata <- contrast_map[contrast_metadata, on = "id"]

# Filter for unique BioSample-level information
keep_cols <- c(
  "BioSample",
  "cell_line",
  "stripped_cell_line",
  "depmap_model_type",
  "oncotree_lineage",
  "oncotree_primary_disease",
  "oncotree_subtype",
  "oncotree_code",
  "sample_collection_site",
  "ccle_name"
)

biosample_metadata <- unique(biosample_metadata[, ..keep_cols])
setDF(biosample_metadata, rownames = biosample_metadata$BioSample)
```

Replace the old colData in the SE object

```{r}
new_coldata <- merge(
  colData(se),
  biosample_metadata,
  by = "BioSample",
  all.x = TRUE
)
rownames(new_coldata) <- new_coldata$BioSample
new_coldata <- new_coldata[colnames(se), ]

# Add onto control Biosample SE object
colData(se) <- NULL
colData(se) <- new_coldata
```

We can also remove any sample from PRIMARY cells and the samples from hTERT cells and samples
not annotated by CCLE.

```{r}
# Samples that were processed but not included in the final db do not have cell line anno
se <- se[, !is.na(se$cell_line)]

# Remove any control samples from primary cells
se <- se[, !se$cell_line %like% "PRIMARY"]

# hTERT samples do not have oncotree lineage
se <- se[, se$oncotree_lineage != ""]

# Samples with CCLE annotations
se <- se[, se$ccle_name != ""]
```

## Dimensionality Reduction on the TEDEdb control samples

### Filter for expression

First, keep only genes and 'main' TE families

```{r}
gtf_gr <- readRDS(
  "/mnt/data/gdata/human/REdiscoverTE_hg38/gencode.v26.annotation.gtf.rds"
)
coding <- gtf_gr[gtf_gr$type == "gene" & gtf_gr$gene_type == "protein_coding", ]
coding <- unique(coding$gene_name)

keep_gene <- rownames(se)[rownames(se) %chin% coding]
keep_te <- rownames(se)[
  rownames(se) %like% "^LINE\\.|^SINE\\.|^LTR\\.|.*\\.SVA\\..*|^DNA\\."
]

keep_feature <- c(keep_gene, keep_te)
```

Now filter for expression based on the primary cancer lineage. Increase the trim for TMM 
normalization to account for cell line differences

```{r}
y <- SE2DGEList(se)
y <- y[rownames(y) %chin% keep_feature, ]

keep <- filterByExpr(y, group = y$samples2$oncotree_lineage)
y <- y[keep, ]

y <- normLibSizes(y, method = "TMM", logratioTrim = 0.5)
lcpms <- cpm(y, log = TRUE)
```

### Perform PCA on genes and TEs separately

Both represent ~80% total variance captured 

```{r}
# ~80% Variance
pca.genes <- PCAtools::pca(
  mat = lcpms[rownames(lcpms) %chin% coding, ],
  metadata = y$samples,
  center = TRUE,
  scale = TRUE,
  removeVar = 0.8,
  rank = 50,
  BSPARAM = BiocSingular::FastAutoParam()
)

# ~80% Variance
pca.tes <- PCAtools::pca(
  mat = lcpms[!rownames(lcpms) %chin% coding, ],
  metadata = y$samples,
  center = TRUE,
  scale = TRUE,
  removeVar = 0.5,
  rank = 100,
  BSPARAM = BiocSingular::FastAutoParam()
)
```

Show PCA biplots

```{r}
b1 <- PCAtools::biplot(
  pca.genes,
  colby = "oncotree_lineage",
  lab = NULL,
  title = "Protein Coding Genes",
  pointSize = 0.8,
  legendPosition = "right"
) +
  labs(color = "OncoTree Lineage") +
  scale_color_paletteer_d("ggsci::default_igv")

b2 <- PCAtools::biplot(
  pca.tes,
  colby = "oncotree_lineage",
  lab = NULL,
  title = "TEs",
  pointSize = 0.8,
  legendPosition = "right"
) +
  labs(color = "OncoTree Lineage") +
  scale_color_paletteer_d("ggsci::default_igv")

(b1 | b2) +
  plot_layout(guides = "collect") &
  guides(color = guide_legend(ncol = 1))
ggsave(here(figures, "pca-biplots.png"), width = 20, height = 8, dpi = 600)
```

Perform UMAP on the PC data

```{r}
udata.genes <- uwot::umap(
  pca.genes$rotated,
  n_neighbors = 50,
  min_dist = 2,
  seed = 42
)
udata.genes <- cbind(data.frame(udata.genes), y$samples)

udata.tes <- uwot::umap(
  pca.tes$rotated,
  n_neighbors = 50,
  min_dist = 2,
  seed = 42
)
udata.tes <- cbind(data.frame(udata.tes), y$samples)
```

Show UMAPs for genes and TEs

```{r}
lin_pal <- c(
  '#e6194b',
  '#3cb44b',
  '#ffe119',
  '#4363d8',
  '#f58231',
  '#911eb4',
  '#46f0f0',
  '#f032e6',
  '#bcf60c',
  '#fabebe',
  '#008080',
  '#e6beff',
  '#9a6324',
  '#fffac8',
  '#800000',
  '#aaffc3',
  '#808000',
  '#000075'
)
names(lin_pal) <- levels(factor(y$samples$oncotree_lineage))

plotUmap <- function(x, colby, color_pal, plot_title = "") {
  plot.new()
  plot.window(xlim = range(x$X1), ylim = range(x$X2))
  grid()
  points(x$X1, x$X2, bg = color_pal[x[[colby]]], pch = 21, cex = 1.5)
  title(xlab = "UMAP1", adj = 0.1, font.lab = 2, cex.lab = 1.5)
  title(ylab = "UMAP2", adj = 0.1, font.lab = 2, cex.lab = 1.5)
  title(main = plot_title, adj = 0, font.main = 2, cex.main = 2)
}

# png(here(figures, "umap_lineage.png"), width = 18, height = 6, units = "in", res = 600)
pdf(here(figures, "umap_lineage.pdf"), width = 18, height = 6)
par(mfrow = c(1, 3))
plotUmap(udata.genes, colby = "oncotree_lineage", lin_pal, "Protein Coding Genes")
plotUmap(udata.tes, colby = "oncotree_lineage", lin_pal, "TEs")

# Add legend as new plot area
plot.new()
plot.window(xlim = c(0, 0.5), ylim = c(1, 10))
legend(
  0,
  10,
  legend = names(lin_pal),
  col = lin_pal,
  pch = 19,
  bty = "n",
  cex = 1.5,
  title = "OncoTree Lineage",
  title.font = 2,
  title.adj = 0
)
dev.off()
```

Color by BioProject

```{r}
prj_pal <- hcl.colors(n = length(unique(y$samples$BioProject)), palette = "viridis")
names(prj_pal) <- levels(factor(y$samples$BioProject))

# png(here(figures, "umap_bioproject.png"), width = 16, height = 8, units = "in", res = 600)
pdf(here(figures, "umap_bioproject.pdf"), width = 16, height = 8)
par(mfrow = c(1, 2))
plotUmap(udata.genes, colby = "BioProject", prj_pal, "Protein Coding Genes")
plotUmap(udata.tes, colby = "BioProject", prj_pal, "TEs")
dev.off()
```

Color by primary disease

```{r}
dis_pal <- as.character(paletteer_d(
  "ggsci::default_igv",
  n = length(unique(y$samples$oncotree_primary_disease))
))
names(dis_pal) <- levels(factor(y$samples$oncotree_primary_disease))

png(here(figures, "umap_disease.png"), width = 18, height = 6, units = "in", res = 600)
par(mfrow = c(1, 3))
plotUmap(udata.genes, "oncotree_primary_disease", dis_pal, "Protein Coding Genes")
plotUmap(udata.tes, "oncotree_primary_disease", dis_pal, "TEs")

# Add legend as new plot area
plot.new()
plot.window(xlim = c(0, 0.5), ylim = c(1, 10))
legend(
  0,
  10,
  legend = names(dis_pal),
  col = dis_pal,
  pch = 19,
  bty = "n",
  cex = 1,
  title = "Primary Disease",
  title.font = 2,
  title.adj = 0
)
dev.off()
```

## Create highlight images to see what the trends are per lineage

```{r}
ggplot(udata.genes, aes(X1, X2, color = oncotree_lineage)) +
  geom_point(size = 0.8) +
  gghighlight::gghighlight(
    use_direct_label = FALSE,
    unhighlighted_params = list(colour = alpha("grey", 0.7))
  ) +
  guides(color = "none") +
  facet_wrap(~oncotree_lineage) +
  labs(x = "UMAP 1", y = "UMAP 2") +
  scale_color_manual(values = lin_pal) +
  coriell::theme_coriell() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank()
  )
ggsave(here(figures, "umap-highlighted_genes.png"), width = 20, height = 12, dpi = 600)

# And for TEs
ggplot(udata.tes, aes(X1, X2, color = oncotree_lineage)) +
  geom_point(size = 0.8) +
  gghighlight::gghighlight(
    use_direct_label = FALSE,
    unhighlighted_params = list(colour = alpha("grey", 0.7))
  ) +
  guides(color = "none") +
  facet_wrap(~oncotree_lineage) +
  labs(x = "UMAP 1", y = "UMAP 2") +
  scale_color_manual(values = lin_pal) +
  coriell::theme_coriell() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank()
  )
ggsave(here(figures, "umap-highlighted_tes.png"), width = 20, height = 12, dpi = 600)
```

## Evaluate performance of TEs and Gene 'clusters'

Here the goal is to test if the variance explained by batch is greater in the TEs than in the genes.
To do this, estimate the variance explained by BioProject and then lineage using PERMANOVA.

```{r}
set.seed(123)

# Compoute euclidean distance between samples
d_genes <- vegan::vegdist(pca.genes$rotated, method = "euclidean")
d_tes <- vegan::vegdist(pca.tes$rotated, method = "euclidean")

# Compute PERMANOVA to get variance explained
BioProject <- pca.genes$metadata$BioProject
Lineage <- pca.genes$metadata$oncotree_lineage

permanova_genes <- vegan::adonis2(
  d_genes ~ Lineage + BioProject,
  by = "terms",
  permutations = 1e3,
  parallel = 48
)

permanova_tes <- vegan::adonis2(
  d_tes ~ Lineage + BioProject,
  by = "terms",
  permutations = 1e3,
  parallel = 48
)
```

Extract and show stats

```{r}
r2_genes <- permanova_genes$R2[1]
p_genes <- permanova_genes$`Pr(>F)`[1]

cat("Gene Expression PERMANOVA:\n")
cat("R-squared:", round(r2_genes, 4), "| p-value:", p_genes, "\n\n")

r2_tes <- permanova_tes$R2[1]
p_tes <- permanova_tes$`Pr(>F)`[1]

cat("TE Expression PERMANOVA:\n")
cat("R-squared:", round(r2_tes, 4), "| p-value:", p_tes, "\n")
```

## Estimate the DSC values

Compute confidence intervals around the dispersion separability criteria value for both datasets 
from: https://bioinformatics.mdanderson.org/public-software/tcga-batch-effects/

```{r}
bootstrap_dsc <- function(x, g, B = 100, conf = 0.95) {
  x <- as.matrix(x)
  g <- as.factor(g)
  observed_dsc <- coriell::dsc(x, g)

  boot_dsc_values <- numeric(B)
  indices <- seq_len(ncol(x))

  strata <- split(indices, g)

  for (i in seq_len(B)) {
    boot_indices <- unlist(lapply(strata, function(idx) {
      sample(idx, size = length(idx), replace = TRUE)
    }))

    x_boot <- x[, boot_indices, drop = FALSE]
    g_boot <- g[boot_indices]

    boot_dsc_values[i] <- coriell::dsc(x_boot, g_boot)
  }

  alpha <- 1 - conf
  lower_prob <- alpha / 2
  upper_prob <- 1 - (alpha / 2)

  ci <- quantile(boot_dsc_values, probs = c(lower_prob, upper_prob), na.rm = TRUE)

  return(list(
    observed = observed_dsc,
    conf_int = ci,
    bootstrap_values = boot_dsc_values
  ))
}

dsc.b <- bootstrap_dsc(lcpms, y$samples$BioProject)
dsc.genes.b <- bootstrap_dsc(lcpms[rownames(lcpms) %chin% coding, ], y$samples$BioProject)
dsc.tes.b <- bootstrap_dsc(lcpms[!(rownames(lcpms) %chin% coding), ], y$samples$BioProject)
```

Evaluate significance

```{r}
permutation_test_dsc <- function(x, g, n_perms = 1000) {
  observed_dsc <- coriell::dsc(x, g)
  null_dsc_values <- numeric(n_perms)

  for (i in seq_len(n_perms)) {
    g_permuted <- sample(g)
    null_dsc_values[i] <- coriell::dsc(x, g_permuted)
  }

  p_value <- (sum(null_dsc_values >= observed_dsc) + 1) / (n_perms + 1)

  return(list(
    observed = observed_dsc,
    p_value = p_value,
    null_distribution = null_dsc_values
  ))
}

dsc.perm <- permutation_test_dsc(lcpms, y$samples$BioProject)
dsc.genes.perm <- permutation_test_dsc(lcpms[rownames(lcpms) %chin% coding, ], y$samples$BioProject)
dsc.tes.perm <- permutation_test_dsc(
  lcpms[!(rownames(lcpms) %chin% coding), ],
  y$samples$BioProject
)
```