---
title: Collect overall database stats
author: Gennaro Calendo
format: html
---

## Set up

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(patchwork))
suppressPackageStartupMessages(library(paletteer))
suppressPackageStartupMessages(library(Ternary))
suppressPackageStartupMessages(library(gt))
suppressPackageStartupMessages(library(glmmTMB))

datafiles <- here("analysis", "results", "data-files", "04")
figures <- here("analysis", "results", "figures", "04")
rdsfiles <- here("analysis", "results", "rds-files", "04")

invisible(sapply(
  c(datafiles, figures, rdsfiles),
  dir.create,
  showWarnings = F,
  recursive = T
))
```

## Create pie charts for overview dashboard

```{r}
# Read in the shell data
se <- readRDS(here("appdata", "se_hdf5", "se.rds"))

# Extract contrast-level data as data.table
rd <- as.data.table(data.frame(colData(se)))

# Get the unique number of contrasts for each variable
experiments <- rd[, .N, by = experiment][order(N, decreasing = TRUE)]
contrasts <- rd[, .N, by = contrast][order(N, decreasing = TRUE)]
cell_lines <- rd[, .N, by = cell_line][order(N, decreasing = TRUE)]
cancer_types <- rd[, .N, by = oncotree_primary_disease][order(
  N,
  decreasing = TRUE
)]
treatments <- rd[, .N, by = treatment][order(N, decreasing = TRUE)]

# Now get the top records for each
top_contrasts <- rd[,
  lapply(.SD, forcats::fct_lump_n, n = 5),
  .SDcols = c("cell_line", "oncotree_primary_disease", "treatment")
]

plot_donut <- function(var, hsize = 2) {
  df <- data.frame(table(top_contrasts[[var]]))

  ggplot(df, aes(x = hsize, y = Freq, fill = Var1)) +
    geom_col(color = "black") +
    geom_text(aes(label = Freq), position = position_stack(vjust = 0.5)) +
    coord_polar(theta = "y") +
    xlim(c(0.2, hsize + 0.5)) +
    theme(
      panel.background = element_rect(fill = "white"),
      panel.grid = element_blank(),
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      axis.text = element_blank(),
      plot.title = element_text(face = "bold", size = 24)
    )
}

p1 <- plot_donut("cell_line") +
  scale_fill_paletteer_d("ggsci::default_nejm") +
  labs(title = "Cell Lines", fill = "Cell Line")

p2 <- plot_donut("oncotree_primary_disease") +
  scale_fill_paletteer_d("ggsci::default_uchicago") +
  labs(title = "Cancer Types", fill = "Cancer Type")

p3 <- plot_donut("treatment") +
  scale_fill_paletteer_d("ggsci::default_jama") +
  labs(title = "Treatments", fill = "Treatment")

(p1 | p2 | p3)
ggsave(here(figures, "records.pdf"), width = 16, height = 8)
ggsave(here(figures, "records.png"), width = 16, height = 8, dpi = 600)
```

## Read in the db data

Now, I need to compute the percentage DE genes/TEs per tissue type, cell line, treatment class, 
etc. to get overall statistics about the database contents.

```{r}
de_datafile <- here(rdsfiles, "de-data.fst")
if (!file.exists(de_datafile)) {
  suppressPackageStartupMessages(library(HDF5Array))
  suppressPackageStartupMessages(library(DelayedArray))
  suppressPackageStartupMessages(library(DelayedMatrixStats))

  db <- loadHDF5SummarizedExperiment("/mnt/data/data_gc/TEDEdb/appdata/se_hdf5")
  getDeDatafromAssays <- function(x) {
    filtered <- db[, x]
    assay_data <- lapply(names(SummarizedExperiment::assays(filtered)), \(x) {
      SummarizedExperiment::assay(filtered, x)
    })

    df <- as.data.frame(do.call(SummarizedExperiment::cbind, assay_data))
    colnames(df) <- names(SummarizedExperiment::assays(filtered))
    data.table::setDT(df, keep.rownames = "feature_id")
    df <- df[!is.na(adj.P.Val), ]
    setorder(df, adj.P.Val)

    return(df)
  }

  de_data <- parallel::mclapply(db$id, getDeDatafromAssays, mc.cores = 32)
  names(de_data) <- db$id
  de_data <- rbindlist(de_data, idcol = "id")
  fst::write_fst(de_data, de_datafile)
} else {
  de_data <- fst::read_fst(de_datafile, as.data.table = TRUE)
}
```

## Read in the shell to get the contrast metadata

```{r}
db <- readRDS(here("appdata", "se_hdf5", "se.rds"))
meta <- setDT(data.frame(colData(db)))
```

## Compute the number of up/down/non-de features per contrast

```{r}
de_data[,
  is_te := feature_id %like% "^LINE\\.|^SINE\\.|^LTR\\.|.*\\.SVA\\..*|^DNA\\."
]

de_stats <- de_data[,
  coriell::summarize_dge(.SD, fdr = 0.1, fdr_col = "adj.P.Val"),
  by = "id"
]
gene_stats <- de_data[
  is_te == FALSE,
  coriell::summarize_dge(.SD, fdr = 0.1, fdr_col = "adj.P.Val"),
  by = "id"
]
te_stats <- de_data[
  is_te == TRUE,
  coriell::summarize_dge(.SD, fdr = 0.1, fdr_col = "adj.P.Val"),
  by = "id"
]
```

Cast wide and add on metadata

```{r}
# Cast stats wider so metadata can be joined onto contrasts
gene_stats.w <- dcast(gene_stats, id ~ Direction, value.var = "N")
te_stats.w <- dcast(te_stats, id ~ Direction, value.var = "N")

# And join on the metadata
gene_stats.w <- meta[gene_stats.w, on = "id"]
te_stats.w <- meta[te_stats.w, on = "id"]

# Remove samples from non-cancer studies
gene_stats.w <- gene_stats.w[
  oncotree_primary_disease != "Non-Cancerous" & !is.na(oncotree_primary_disease)
]
te_stats.w <- te_stats.w[
  oncotree_primary_disease != "Non-Cancerous" & !is.na(oncotree_primary_disease)
]

# Remove any samples from primary or multiple cell lines
gene_stats.w <- gene_stats.w[!cell_line %like% "^PRIMARY|MULTIPLE"]
te_stats.w <- te_stats.w[!cell_line %like% "^PRIMARY|MULTIPLE"]

# # Remove non-compound treatments
drop_mech <- c(
  "N/A",
  "small interfering rna",
  "short hairpin rna",
  "hyperthermia",
  "cas9 editing",
  "control vehicle",
  "lncrna targeting astin1",
  "lncrna targeting nkx2",
  "protac (nonfunctional)",
  "radiation therapy",
  "radiation therapy | topoisomerase inhibitor",
  "sparcle expression",
  "sparcle expression regulator",
  "sparcle expression regulator"
)
gene_stats.w <- gene_stats.w[
  clinical_phase != "N/A" & (!mechanism_of_action %chin% drop_mech)
]
te_stats.w <- te_stats.w[
  clinical_phase != "N/A" & (!mechanism_of_action %chin% drop_mech)
]

# # Drop any contrasts that have been flagged as having outlier samples
gene_stats.w <- gene_stats.w[outlier_flags == "None"]
te_stats.w <- te_stats.w[outlier_flags == "None"]

# Also remove 'Compound Kushen Injection' since these are technically combintaions
gene_stats.w <- gene_stats.w[!treatment %like% 'Compound Kushen Injection']
te_stats.w <- te_stats.w[!treatment %like% 'Compound Kushen Injection']

# Add DE percentasge
gene_stats.w[, `:=`(
  prop_up = Up / (Up + Down + Unperturbed),
  prop_down = Down / (Up + Down + Unperturbed),
  prop_non = Unperturbed / (Up + Down + Unperturbed)
)]

te_stats.w[, `:=`(
  prop_up = Up / (Up + Down + Unperturbed),
  prop_down = Down / (Up + Down + Unperturbed),
  prop_non = Unperturbed / (Up + Down + Unperturbed)
)]
```

### Show overall stats

Show a ternary plot with the surface colored by the density of percentage dysregulation

```{r}
# pdf(here(figures, "contrast-ternary.pdf"), width = 12, height = 6)
png(
  here(figures, "contrast-ternary.png"),
  width = 12,
  height = 6,
  units = "in",
  res = 600
)

par(mfrow = c(1, 2))

# Genes
TernaryPlot(
  alab = "Up-Regulated (%)",
  blab = "Down-Regulated (%)",
  clab = "Unchanged (%)",
  point = "up",
  lab.cex = 1.2,
  grid.lines = 10,
  grid.minor.lines = 0,
  grid.lty = 3,
  main = "Protein Coding Genes",
  cex.main = 1.5
)

gene_coords <- gene_stats.w[, .(prop_up, prop_down, prop_non)]
gene_density <- TernaryDensity(gene_coords, resolution = 30L)

# Convert density to absolute numbers
gene_total_dens <- sum(gene_density["z", ], na.rm = TRUE)
gene_frac <- gene_density["z", ] / gene_total_dens
gene_counts <- gene_frac * nrow(gene_stats.w)

# Add back to matrix
gene_density["z", ] <- round(log10(gene_counts + 1), 1)

ColourTernary(
  gene_density,
  spectrum = hcl.colors(30, palette = "Viridis", rev = TRUE),
  legend = c('1,000', '100', '10', '0'),
  bty = "n",
  title = "Number of Contrasts",
  cex = 0.8
)

# TEs
TernaryPlot(
  alab = "Up-Regulated (%)",
  blab = "Down-Regulated (%)",
  clab = "Unchanged (%)",
  point = "up",
  lab.cex = 1.2,
  grid.lines = 10,
  grid.minor.lines = 0,
  grid.lty = 3,
  main = "TE Subfamilies",
  cex.main = 1.5
)

te_coords <- te_stats.w[, .(prop_up, prop_down, prop_non)]
te_density <- TernaryDensity(te_coords, resolution = 30L)

# Convert density to absolute numbers
te_total_dens <- sum(te_density["z", ], na.rm = TRUE)
te_frac <- te_density["z", ] / te_total_dens
te_counts <- te_frac * nrow(te_stats.w)

# Add back to matrix
te_density["z", ] <- round(log10(te_counts + 1), 1)

ColourTernary(
  te_density,
  spectrum = hcl.colors(30, palette = "Viridis", rev = TRUE),
  legend = c('1,000', '100', '10', '0'),
  bty = "n",
  title = "Number of Contrasts",
  cex = 0.8
)

dev.off()
```

## Dysregulation by epigenetic class

```{r}
gene_by_epi <- gene_stats.w[,
  .(
    avg_prop_up = mean(prop_up),
    avg_prop_down = mean(prop_down),
    sem_prop_up = sd(prop_up) / sqrt(.N),
    sem_prop_down = sd(prop_down) / sqrt(.N),
    N = .N
  ),
  by = epigenetic_class_fine
]

te_by_epi <- te_stats.w[,
  .(
    avg_prop_up = mean(prop_up),
    avg_prop_down = mean(prop_down),
    sem_prop_up = sd(prop_up) / sqrt(.N),
    sem_prop_down = sd(prop_down) / sqrt(.N),
    N = .N
  ),
  by = epigenetic_class_fine
]

# Melt longer for plotting
gene_by_epi1 <- melt(
  gene_by_epi,
  id.vars = c("epigenetic_class_fine", "N"),
  measure.vars = c("avg_prop_up", "avg_prop_down"),
  variable.name = "direction",
  value.name = "avg"
)
gene_by_epi2 <- melt(
  gene_by_epi,
  id.vars = "epigenetic_class_fine",
  measure.vars = c("sem_prop_up", "sem_prop_down"),
  variable.name = "direction",
  value.name = "sem"
)
gene_by_epi <- cbind(gene_by_epi1, gene_by_epi2[, .(sem)])

# And for TEs
te_by_epi1 <- melt(
  te_by_epi,
  id.vars = c("epigenetic_class_fine", "N"),
  measure.vars = c("avg_prop_up", "avg_prop_down"),
  variable.name = "direction",
  value.name = "avg"
)
te_by_epi2 <- melt(
  te_by_epi,
  id.vars = "epigenetic_class_fine",
  measure.vars = c("sem_prop_up", "sem_prop_down"),
  variable.name = "direction",
  value.name = "sem"
)
te_by_epi <- cbind(te_by_epi1, te_by_epi2[, .(sem)])

# Keep only the single classes
keep_classes <- c("DNMTi", "HMTi", "Non-epigenetic", "HDMi", "HDACi", "PAHi")
gene_by_epi <- gene_by_epi[epigenetic_class_fine %chin% keep_classes]
te_by_epi <- te_by_epi[epigenetic_class_fine %chin% keep_classes]

# Invert the down percentages
gene_by_epi[direction %like% "down", avg := avg * -1]
te_by_epi[direction %like% "down", avg := avg * -1]

# Swap labels for plotting
gene_by_epi[,
  direction := fifelse(direction %like% "up", "Up-regulated", "Down-regulated")
]
te_by_epi[,
  direction := fifelse(direction %like% "up", "Up-regulated", "Down-regulated")
]

# Create new x-axis labels
gene_by_epi[, xlabel := paste0(epigenetic_class_fine, "\n(N=", N, ")")]
te_by_epi[, xlabel := paste0(epigenetic_class_fine, "\n(N=", N, ")")]

# Reorder factor levels
gene_by_epi[,
  xlabel := factor(
    xlabel,
    levels = c(
      "DNMTi\n(N=57)",
      "HMTi\n(N=26)",
      "Non-epigenetic\n(N=1163)",
      "HDMi\n(N=9)",
      "HDACi\n(N=62)",
      "PAHi\n(N=70)"
    )
  )
]
te_by_epi[,
  xlabel := factor(
    xlabel,
    levels = c(
      "DNMTi\n(N=57)",
      "HMTi\n(N=26)",
      "Non-epigenetic\n(N=1163)",
      "HDMi\n(N=9)",
      "HDACi\n(N=62)",
      "PAHi\n(N=70)"
    )
  )
]
```

Show the hanging barplots

```{r}
# Genes
p1 <- ggplot(gene_by_epi, aes(x = xlabel, y = avg, fill = direction)) +
  geom_col(color = "white", width = 0.8) +
  geom_errorbar(aes(ymin = avg - sem, ymax = avg + sem, width = 0.2)) +
  scale_fill_manual(
    values = c("Down-regulated" = "steelblue", "Up-regulated" = "firebrick2")
  ) +
  scale_y_continuous(
    labels = \(x) paste0(abs(x * 100), "%"),
    limits = c(-0.2, 0.2)
  ) +
  labs(
    title = "Protein Coding Genes",
    y = "Percentage Dysregulation (mean \u00B1 SEM)",
    x = NULL,
    fill = NULL
  ) +
  see::theme_radar() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "lightgrey", linetype = 2),
    plot.title = element_text(size = 18, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# TEs
p2 <- ggplot(te_by_epi, aes(x = xlabel, y = avg, fill = direction)) +
  geom_col(color = "white", width = 0.8) +
  geom_errorbar(aes(ymin = avg - sem, ymax = avg + sem, width = 0.2)) +
  scale_fill_manual(
    values = c("Down-regulated" = "steelblue", "Up-regulated" = "firebrick2")
  ) +
  scale_y_continuous(
    labels = \(x) paste0(abs(x * 100), "%"),
    limits = c(-0.2, 0.2)
  ) +
  labs(
    title = "TE Subfamilies",
    y = "Percentage Dysregulation (mean \u00B1 SEM)",
    x = NULL,
    fill = NULL
  ) +
  see::theme_radar() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "lightgrey", linetype = 2),
    plot.title = element_text(size = 18, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

(p1 | p2) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
ggsave(
  here(figures, "dysregulation-barplot.png"),
  width = 11,
  height = 6,
  dpi = 600
)
ggsave(here(figures, "dysregulation-barplot.pdf"), width = 11, height = 6)
```

## For each epigenetic class, what are the top *compounds* by percentage TE dysregulation?

- Average gene and TE dysregulation by compound
- Select the top 5-10 from each 
- Crate a table

```{r}
te_by_compound <- te_stats.w[
  epigenetic_class_fine %chin% keep_classes,
  .(
    avg_prop_non = mean(prop_non, na.rm = TRUE),
    avg_prop_up = mean(prop_up, na.rm = TRUE),
    avg_prop_down = mean(prop_down, na.rm = TRUE),
    sd_prop_up = sd(prop_up, na.rm = TRUE),
    sd_prop_down = sd(prop_down, na.rm = TRUE),
    N = .N,
    cell_lines = paste(unique(cell_line), collapse = "; "),
    n_cell_lines = length(unique(cell_line)),
    lineages = paste(unique(oncotree_lineage), collapse = "; "),
    bioprojects = paste(unique(experiment), collapse = "; "),
    n_bioprojects = length(unique(experiment))
  ),
  by = .(epigenetic_class_fine, treatment, mechanism_of_action)
]
```

### Perform shrinkage on mean estimates to get robust results and CIs

```{r}
fit.up <- glmmTMB(
  prop_up ~ 1 + (1 | epigenetic_class_fine / treatment),
  data = te_stats.w,
  family = tweedie(link = "log")
)

fit.down <- glmmTMB(
  prop_down ~ 1 + (1 | epigenetic_class_fine / treatment),
  data = te_stats.w,
  family = tweedie(link = "log")
)

# Extract the tested compounds
shrunken_dt <- data.table(trt_class = rownames(coef(fit.up)[[1]][[1]]))
shrunken_dt[, c("treatment", "epigenetic_class_fine") := tstrsplit(trt_class, ":", fixed = TRUE)][,
  trt_class := NULL
]

# Predict the shrunken means
pred_up <- predict(fit.up, newdata = shrunken_dt, type = "link", se.fit = TRUE)
pred_down <- predict(fit.down, newdata = shrunken_dt, type = "link", se.fit = TRUE)

# Compute 95% CIs
up.lower <- pred_up$fit - (1.96 * pred_up$se.fit)
up.upper <- pred_up$fit + (1.96 * pred_up$se.fit)
down.lower <- pred_down$fit - (1.96 * pred_down$se.fit)
down.upper <- pred_down$fit + (1.96 * pred_down$se.fit)

# Exponentiate back to percentage scale
shrunken_dt[, `:=`(
  predicted_up = exp(pred_up$fit),
  up_lower = exp(up.lower),
  up_upper = exp(up.upper),
  predicted_down = exp(pred_down$fit),
  down_lower = exp(down.lower),
  down_upper = exp(down.upper)
)]
```

Join the shrunken data onto the raw data

```{r}
merged_by_compound <- merge(
  x = te_by_compound,
  y = shrunken_dt,
  by = c("epigenetic_class_fine", "treatment"),
  all.x = TRUE
)
setorder(merged_by_compound, epigenetic_class_fine, -predicted_up)

# Limit to compounds observed in >1 BioProjects
top_te_by_compound <- merged_by_compound[n_bioprojects > 2]
```

Create the final table. For now, gt_save() is not producing a nice table. Open in browser and save 
as pdf.

```{r}
keep_cols <- c(
  "treatment",
  "predicted_up",
  "up_lower",
  "up_upper",
  "predicted_down",
  "down_lower",
  "down_upper",
  "N",
  "n_cell_lines",
  "n_bioprojects",
  "mechanism_of_action"
)

format_ci <- function(est, lower, upper, digits = 1) {
  paste0(
    round(est * 100, digits),
    "% [",
    round(lower * 100, digits),
    " - ",
    round(upper * 100, digits),
    "]%"
  )
}

# Construct gt table
tbl <- top_te_by_compound[, ..keep_cols]
setorder(tbl, -predicted_up)
tbl[, avg_prop_up := format_ci(predicted_up, up_lower, up_upper)]
tbl[, avg_prop_down := format_ci(predicted_down, down_lower, down_upper)]
tbl[, `:=`(
  predicted_up = NULL,
  predicted_down = NULL,
  up_upper = NULL,
  up_lower = NULL,
  down_upper = NULL,
  down_lower = NULL
)]
setcolorder(
  tbl,
  c(
    "treatment",
    "mechanism_of_action",
    "avg_prop_up",
    "avg_prop_down",
    "N",
    "n_bioprojects",
    "n_cell_lines"
  )
)

# Construct final figure
gt(tbl) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  cols_align(
    align = "right",
    columns = c(avg_prop_up, avg_prop_down)
  ) |>
  cols_align(
    align = "center",
    columns = c(N, n_cell_lines, n_bioprojects, mechanism_of_action)
  ) |>
  cols_align(
    align = "left",
    columns = c(treatment)
  ) |>
  cols_label(
    mechanism_of_action = "Mechanism of Action",
    treatment = "Compound",
    avg_prop_up = "Up-regulation [95% CI]",
    avg_prop_down = "Down-regulation [95% CI]",
    N = "Contrasts (N)",
    n_cell_lines = "Cell Lines (N)",
    n_bioprojects = "BioProjects (N)"
  ) |>
  cols_width(
    'treatment' ~ px(150),
    'N' ~ px(150),
    'n_cell_lines' ~ px(150),
    'n_bioprojects' ~ px(150)
  ) |>
  tab_style(
    style = cell_fill(color = "#FFFFE0"),
    locations = cells_body(rows = mechanism_of_action %like% "cdk inhibitor")
  ) |>
  opt_stylize(style = 1, color = "gray", add_row_striping = FALSE)
```

## Top dysregulated *features* in each class

### Connect to TEDEdb

```{r}
db <- HDF5Array::loadHDF5SummarizedExperiment("/mnt/data/data_gc/TEDEdb/appdata/se_hdf5")
```

### Filter contrasts

Same as above

```{r}
# Remove samples from non-cancer studies
db <- db[, db$oncotree_primary_disease != "Non-Cancerous" & !is.na(db$oncotree_primary_disease)]

# Remove any samples from primary or multiple cell lines
db <- db[, !db$cell_line %like% "^PRIMARY|MULTIPLE"]

# Remove non-compound treatments
drop_mech <- c(
  "N/A",
  "small interfering rna",
  "short hairpin rna",
  "hyperthermia",
  "cas9 editing",
  "control vehicle",
  "lncrna targeting astin1",
  "lncrna targeting nkx2",
  "protac (nonfunctional)",
  "radiation therapy",
  "radiation therapy | topoisomerase inhibitor",
  "sparcle expression",
  "sparcle expression regulator",
  "sparcle expression regulator"
)
db <- db[, db$clinical_phase != "N/A" & (!db$mechanism_of_action %chin% drop_mech)]

# # Drop any contrasts that have been flagged as having outlier samples
db <- db[, db$outlier_flags == "None"]

# Also remove 'Compound Kushen Injection' since these are technically combintaions
db <- db[, !db$treatment %like% 'Compound Kushen Injection']
```

### Determine top dysregulated features

Compute the frequency of dysregulation across contrasts per class

```{r}
deFrequency <- function(x, mechanism = NULL) {
  if (!is.null(mechanism)) {
    se <- db[, db$mechanism_of_action == mechanism]
  } else {
    se <- db[, db$epigenetic_class_fine == x]
  }

  is_up <- (assay(se, "logFC") > 0) & (assay(se, "adj.P.Val") < 0.1)
  is_down <- (assay(se, "logFC") < 0) & (assay(se, "adj.P.Val") < 0.1)

  result <- data.table(
    feature = rownames(se),
    n_contrasts = ncol(se),
    n_up = rowSums(is_up),
    n_down = rowSums(is_down)
  )

  result[, n_de := n_up + n_down][, `:=`(
    percentage_up = n_up / n_contrasts * 100,
    percentage_down = n_down / n_contrasts * 100,
    direction_score = (n_up - n_down) / n_de
  )]

  result[is.na(direction_score), direction_score == 0.0]

  # Base percentage DE on direction of DE
  result[,
    percentage_de := fcase(
      direction_score > 0 , n_up / n_contrasts * 100   ,
      direction_score < 0 , n_down / n_contrasts * 100 ,
      default = 0.0
    )
  ]

  return(result)
}

# For each epigenetic class
freq_results1 <- lapply(keep_classes, deFrequency)
names(freq_results1) <- keep_classes

# For CDK inhibitors alone
freq_results2 <- list("CDKi" = deFrequency("", mechanism = "cdk inhibitor"))

# Combine into single DT
freq_results <- c(freq_results1, freq_results2)
freq_results_dt <- rbindlist(freq_results, idcol = "epigenetic_class_fine")
fwrite(freq_results_dt, here(datafiles, "dysregulation-frequencies.tsv"), sep = "\t")
```

Select the top 10 genes in each subset

```{r}
lvls <- c("CDKi", keep_classes)

top_genes <- freq_results_dt[
  !feature %like% ".*\\..*\\..*",
  head(.SD[order(percentage_de, decreasing = T)], n = 10),
  by = epigenetic_class_fine
]

# Get the top genes in each set to highlight on the plot
top_genes_per_class <- lapply(split(top_genes, by = "epigenetic_class_fine"), `[[`, "feature")
top_genes_per_class <- top_genes_per_class[lvls]

# Collect all genes to plot
top_genes_dt <- freq_results_dt[feature %chin% top_genes[, unique(feature)]]

# Reorder to match other plots
top_genes_dt[, epigenetic_class_fine := factor(epigenetic_class_fine, levels = lvls)]
```

Repeat for TEs

```{r}
top_tes <- freq_results_dt[
  feature %like% ".*\\..*\\..*",
  head(.SD[order(percentage_de, decreasing = T)], n = 10),
  by = epigenetic_class_fine
]

# Get the top TEs in each set
top_tes_per_class <- lapply(split(top_tes, by = "epigenetic_class_fine"), `[[`, "feature")
top_tes_per_class <- top_tes_per_class[lvls]

# Collect all TEs to plot
top_tes_dt <- freq_results_dt[feature %chin% top_tes[, unique(feature)]]

# Relevel to match other plots
top_tes_dt[, epigenetic_class_fine := factor(epigenetic_class_fine, levels = lvls)]
```

Determine factor levels of the genes/TEs to reorder by. 

```{r}
gene_order <- unique(unname(unlist(top_genes_per_class)))
te_order <- unique(unname(unlist(top_tes_per_class)))
top_genes_dt[, feature := factor(feature, levels = gene_order)]
top_tes_dt[, feature := factor(feature, levels = te_order)]
```

Create a frequency dotplot of the top genes

```{r}
p1 <- ggplot(top_genes_dt, aes(x = feature, y = epigenetic_class_fine)) +
  geom_point(aes(size = percentage_de, color = direction_score), shape = 1) +
  geom_point(
    data = top_genes_dt[
      epigenetic_class_fine == "CDKi" & feature %chin% top_genes_per_class[["CDKi"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  geom_point(
    data = top_genes_dt[
      epigenetic_class_fine == "DNMTi" & feature %chin% top_genes_per_class[["DNMTi"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  geom_point(
    data = top_genes_dt[
      epigenetic_class_fine == "HDACi" & feature %chin% top_genes_per_class[["HDACi"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  geom_point(
    data = top_genes_dt[
      epigenetic_class_fine == "HDMi" & feature %chin% top_genes_per_class[["HDMi"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  geom_point(
    data = top_genes_dt[
      epigenetic_class_fine == "HMTi" & feature %chin% top_genes_per_class[["HMTi"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  geom_point(
    data = top_genes_dt[
      epigenetic_class_fine == "Non-epigenetic" &
        feature %chin% top_genes_per_class[["Non-epigenetic"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  geom_point(
    data = top_genes_dt[
      epigenetic_class_fine == "PAHi" & feature %chin% top_genes_per_class[["PAHi"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  scale_color_gradient2(
    low = "#313695",
    mid = "#e0f3f8",
    high = "#a50026",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  scale_size_continuous(
    limits = c(0, 100),
    range = c(1, 8),
    breaks = c(0, 25, 50, 75, 100),
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(color = "black", size = 22),
    axis.text.x = element_text(angle = 45, size = 12, hjust = 1, vjust = 1, color = "black"),
    axis.text.y = element_text(color = "black", size = 12),
    panel.grid.major = element_line(color = "grey95"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(
    title = "Protein Coding Genes",
    x = NULL,
    y = NULL,
    color = "Direction Score",
    size = "Percentage DE"
  ) +
  coord_fixed(ratio = 0.90)
```

Create the same plot for TEs

```{r}
p2 <- ggplot(top_tes_dt, aes(x = feature, y = epigenetic_class_fine)) +
  geom_point(aes(size = percentage_de, color = direction_score), shape = 1) +
  geom_point(
    data = top_tes_dt[
      epigenetic_class_fine == "CDKi" & feature %chin% top_tes_per_class[["CDKi"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  geom_point(
    data = top_tes_dt[
      epigenetic_class_fine == "DNMTi" & feature %chin% top_tes_per_class[["DNMTi"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  geom_point(
    data = top_tes_dt[
      epigenetic_class_fine == "HDACi" & feature %chin% top_tes_per_class[["HDACi"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  geom_point(
    data = top_tes_dt[
      epigenetic_class_fine == "HDMi" & feature %chin% top_tes_per_class[["HDMi"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  geom_point(
    data = top_tes_dt[
      epigenetic_class_fine == "HMTi" & feature %chin% top_tes_per_class[["HMTi"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  geom_point(
    data = top_tes_dt[
      epigenetic_class_fine == "Non-epigenetic" &
        feature %chin% top_tes_per_class[["Non-epigenetic"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  geom_point(
    data = top_tes_dt[
      epigenetic_class_fine == "PAHi" & feature %chin% top_tes_per_class[["PAHi"]]
    ],
    aes(size = percentage_de, color = direction_score),
    shape = 16
  ) +
  scale_color_gradient2(
    low = "#313695",
    mid = "#e0f3f8",
    high = "#a50026",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  scale_size_continuous(
    limits = c(0, 100),
    range = c(1, 8),
    breaks = c(0, 25, 50, 75, 100),
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(color = "black", size = 22),
    axis.text.x = element_text(angle = 45, size = 12, hjust = 1, vjust = 1, color = "black"),
    axis.text.y = element_text(color = "black", size = 12),
    panel.grid.major = element_line(color = "grey95"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(
    title = "TE Subfamilies",
    x = NULL,
    y = NULL,
    color = "Direction Score",
    size = "Percentage DE"
  ) +
  coord_fixed(ratio = 0.9)
```

```{r}
(p1 / p2) + plot_layout(guides = "collect")
ggsave(here(figures, "top-features.png"), width = 18, height = 12, dpi = 600)
ggsave(here(figures, "top-features.pdf"), width = 18, height = 12)
```

