---
author: Gennaro Calendo
title: Perform meta-analysis on collections of DE experiments
format: html
---

## Set up

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(coriell))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(patchwork))
suppressPackageStartupMessages(library(legendry))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(gt))
suppressPackageStartupMessages(library(PCAtools))
suppressPackageStartupMessages(library(variancePartition))

datafiles <- here("analysis", "results", "data-files", "05")
figures <- here("analysis", "results", "figures", "05")
rdsfiles <- here("analysis", "results", "rds-files", "05")

invisible(sapply(c(datafiles, figures, rdsfiles), dir.create, showWarnings = F))
```

## Connect to TEDEdb

```{r}
db <- HDF5Array::loadHDF5SummarizedExperiment(
  "/mnt/data/data_gc/TEDEdb/appdata/se_hdf5"
)
```

## Filter contrasts

- Remove any contrasts from primary cells or multiple cell lines combined
- Remove any contrasts from non-cancer cell lines
- Remove any contrasts containing samples flagged as outliers

```{r}
db <- db[, !db$cell_line %like% "^PRIMARY|^MULTIPLE"]
db <- db[,
  db$oncotree_primary_disease != "Non-Cancerous" &
    !is.na(db$oncotree_primary_disease)
]
db <- db[, db$outlier_flags == "None"]
```

## Perform meta-analysis on top contrasts

Pick treatments with the greatest number of contrasts in the database

```{r}
# Take the top treatments by number of contrasts
# head(sort(table(db$treatment), decreasing=T), n=10)

trts <- c(
  "OTS964",
  "Doxorubicin",
  "Decitabine",
  "Actinomycin D",
  "Etoposide",
  "Azacitidine",
  "JQ1",
  "Gemcitabine",
  "THZ1"
)

# How many BioProjects, cell lines, etc. are represented?
for (trt in trts) {
  df <- colData(db)[db$treatment == trt, ]
  msg <- paste0(
    trt,
    " - ",
    "Contrasts: ",
    nrow(df),
    "; ",
    "BioProjects: ",
    length(unique(df$experiment)),
    "; ",
    "Cell Lines: ",
    length(unique(df$cell_line)),
    "; ",
    "Lineages: ",
    length(unique(df$oncotree_lineage)),
    "\n"
  )
  cat(msg)
}
```

For the main:

> Decitabine - Contrasts: 35; BioProjects: 13; Cell Lines: 21; Lineages: 7
> Doxorubicin - Contrasts: 38; BioProjects: 10; Cell Lines: 16; Lineages: 7
> Etoposide - Contrasts: 30; BioProjects: 7; Cell Lines: 25; Lineages: 8
> JQ1 - Contrasts: 20; BioProjects: 7; Cell Lines: 15; Lineages: 5

For the supplement(?):

> Azacitidine - Contrasts: 22; BioProjects: 4; Cell Lines: 10; Lineages: 3
> Gemcitabine - Contrasts: 18; BioProjects: 4; Cell Lines: 5; Lineages: 3
> Actinomycin D - Contrasts: 32; BioProjects: 3; Cell Lines: 3; Lineages: 2
> THZ1 - Contrasts: 16; BioProjects: 4; Cell Lines: 7; Lineages: 3

Exclude since only one BioProject:

> OTS964 - Contrasts: 40; BioProjects: 1; Cell Lines: 5; Lineages: 4

Perform the meta-analysis on these treatments

```{r}
meta_results <- parallel::mclapply(
  trts,
  function(x) {
    se <- db[, db$treatment == x]
    result <- meta_de(
      se,
      FUN = metapod::parallelFisher,
      pval = "P.Value",
      se = "SE"
    )
    result <- result[!is.na(Meta.Pval)]
    result[, Meta.FDR := p.adjust(Meta.Pval, method = "BH")]
    return(result)
  },
  mc.cores = 4
)
names(meta_results) <- trts

# Save results to file
meta_results_dt <- rbindlist(meta_results, idcol = "treatment")
fwrite(
  meta_results_dt,
  here(datafiles, "meta-analysis-results.tsv"),
  sep = "\t"
)
```

### Perform jackknife resampling on the meta-results to assess robustness

```{r}
run <- FALSE

if (run) {
  metafun <- function(x) {
    meta_de(x, metapod::parallelFisher, pval = "P.Value", se = "SE")
  }

  message("Running Decitabine jacknife meta-analysis...")
  dac_jk_results <- jackknifeSE(db[, db$treatment == "Decitabine"], FUN = metafun)
  dac_jk_results <- rbindlist(dac_jk_results, idcol = "Jackknife")
  dac_jk_results <- dac_jk_results[!is.na(Meta.Pval)]
  dac_jk_results[, Meta.FDR := p.adjust(Meta.Pval, method = "BH"), by = Jackknife]
  fwrite(dac_jk_results, here(datafiles, "dac-jackknife-results.tsv"), sep = "\t")

  message("Running Doxorubicin jacknife meta-analysis...")
  dox_jk_results <- jackknifeSE(db[, db$treatment == "Doxorubicin"], FUN = metafun)
  dox_jk_results <- rbindlist(dox_jk_results, idcol = "Jackknife")
  dox_jk_results <- dox_jk_results[!is.na(Meta.Pval)]
  dox_jk_results[, Meta.FDR := p.adjust(Meta.Pval, method = "BH"), by = Jackknife]
  fwrite(dox_jk_results, here(datafiles, "dox-jackknife-results.tsv"), sep = "\t")

  message("Running Etoposide jacknife meta-analysis...")
  eto_jk_results <- jackknifeSE(db[, db$treatment == "Etoposide"], FUN = metafun)
  eto_jk_results <- rbindlist(eto_jk_results, idcol = "Jackknife")
  eto_jk_results <- eto_jk_results[!is.na(Meta.Pval)]
  eto_jk_results[, Meta.FDR := p.adjust(Meta.Pval, method = "BH"), by = Jackknife]
  fwrite(eto_jk_results, here(datafiles, "eto-jackknife-results.tsv"), sep = "\t")

  message("Running JQ1 jacknife meta-analysis...")
  jq1_jk_results <- jackknifeSE(db[, db$treatment == "JQ1"], FUN = metafun)
  jq1_jk_results <- rbindlist(jq1_jk_results, idcol = "Jackknife")
  jq1_jk_results <- jq1_jk_results[!is.na(Meta.Pval)]
  jq1_jk_results[, Meta.FDR := p.adjust(Meta.Pval, method = "BH"), by = Jackknife]
  fwrite(jq1_jk_results, here(datafiles, "jq1-jackknife-results.tsv"), sep = "\t")
} else {
  dac_jk_results <- fread(here(datafiles, "dac-jackknife-results.tsv"))
  dox_jk_results <- fread(here(datafiles, "dox-jackknife-results.tsv"))
  eto_jk_results <- fread(here(datafiles, "eto-jackknife-results.tsv"))
  jq1_jk_results <- fread(here(datafiles, "jq1-jackknife-results.tsv"))
}
```

Compute the jackknife bias and standard error on the percentage differential expression

```{r}
computeJackknifeBias <- function(orig, jacks, n, feature = "genes") {
  if (feature == "genes") {
    dt1 <- orig[!Feature %like% ".*\\..*\\..*"]
    dt2 <- jacks[!Feature %like% ".*\\..*\\..*"]
  } else {
    dt1 <- orig[Feature %like% ".*\\..*\\..*"]
    dt2 <- jacks[Feature %like% ".*\\..*\\..*"]
  }

  # Get the observed data from the full test
  summary_stats <- summarize_dge(
    dt1,
    fdr = 0.1,
    lfc = log2(1.5),
    fdr_col = "Meta.FDR",
    lfc_col = "Meta.logFC"
  )
  setDT(summary_stats)
  theta_hat_up <- summary_stats[Direction == "Up", Percent]
  theta_hat_down <- summary_stats[Direction == "Down", Percent]

  # Compute mean percentages across all jackknifes
  jack_stats <- dt2[,
    summarize_dge(.SD, fdr = 0.1, lfc = log2(1.5), fdr_col = "Meta.FDR", lfc_col = "Meta.logFC"),
    by = Jackknife
  ]

  jack_estimates_up <- jack_stats[Direction == "Up", Percent]
  jack_estimates_down <- jack_stats[Direction == "Down", Percent]
  theta_bar_up <- mean(jack_estimates_up)
  theta_bar_down <- mean(jack_estimates_down)

  # Compute jackknife bias and variance
  bias_jack_up <- (n - 1) * (theta_bar_up - theta_hat_up)
  bias_jack_down <- (n - 1) * (theta_bar_down - theta_hat_down)

  var_term_up <- sum((jack_estimates_up - theta_bar_up)^2)
  var_term_down <- sum((jack_estimates_down - theta_bar_down)^2)

  se_jack_up <- sqrt(((n - 1) / n) * var_term_up)
  se_jack_down <- sqrt(((n - 1) / n) * var_term_down)

  data.table(
    Direction = c("Up", "Down"),
    Observed = c(theta_hat_up, theta_hat_down),
    Corrected = c(theta_hat_up - bias_jack_up, theta_hat_down - bias_jack_down),
    Bias = c(bias_jack_up, bias_jack_down),
    SE = c(se_jack_up, se_jack_down)
  )
}
```

Compute the bias and variance for each set of results

```{r}
jacks <- list(
  Decitabine = dac_jk_results,
  Doxorubicin = dox_jk_results,
  Etoposide = eto_jk_results,
  JQ1 = jq1_jk_results
)

ns <- list(
  Decitabine = ncol(db[, db$treatment == "Decitabine"]),
  Doxorubicin = ncol(db[, db$treatment == "Doxorubicin"]),
  Etoposide = ncol(db[, db$treatment == "Etoposide"]),
  JQ1 = ncol(db[, db$treatment == "JQ1"])
)

origs <- meta_results[names(jacks)]

# Compute jackknife bias and variance for gene percentages
gene_jack_stats <- Map(
  computeJackknifeBias,
  orig = origs,
  jacks = jacks,
  n = ns
)
names(gene_jack_stats) <- paste0(names(gene_jack_stats), " (n=", unlist(ns), ")")
gene_jack_stats <- rbindlist(gene_jack_stats, idcol = "Compound")

# Compute jackknife bias and variance for TE percentages
te_jack_stats <- Map(
  computeJackknifeBias,
  orig = origs,
  jacks = jacks,
  n = ns,
  feature = "TE"
)
names(te_jack_stats) <- paste0(names(te_jack_stats), " (n=", unlist(ns), ")")
te_jack_stats <- rbindlist(te_jack_stats, idcol = "Compound")
```

### Permute the features in each matrix to get baseline for percentages that are expected

JP wanted to know some kind of 'baseline' expectation for how many genes are expected to be 
dysregulated in a given meta-analysis. He suggested permuting the genes and re-computing the 
meta-analysis. My expectation is that this will yield hardly any DE results and will not serve as a 
more useful assesment of stability than the jackknife.

First, create a funtion that will shuffle the entries in each column of the given se object, keeping 
each column self-contained (so all results from a single experiment are shuffled) but synchronized 
across P.value, SE, and logFC matrices. This requires first generating ncol indeces from 1:nrow and 
then applying these shuffled indeces to each matrix of the SE to ensure they stay in sync.  

```{r}
shuffleSE <- function(x) {
  # Realize original matrices in-memory
  orig_lfc <- as.matrix(assay(x, "logFC"))
  orig_pval <- as.matrix(assay(x, "P.Value"))
  orig_stderr <- as.matrix(assay(x, "SE"))

  # Allocate new space for shuffled results
  lfc <- matrix(data = NA_real_, ncol = ncol(x), nrow = nrow(x))
  pval <- matrix(data = NA_real_, ncol = ncol(x), nrow = nrow(x))
  stderr <- matrix(data = NA_real_, ncol = ncol(x), nrow = nrow(x))
  dimnames(lfc) <- dimnames(pval) <- dimnames(stderr) <- dimnames(orig_lfc)

  # Create random row indeces per column and populate matrices
  rand_row_per_col <- replicate(ncol(x), sample.int(nrow(x)), simplify = FALSE)
  for (j in seq_along(rand_row_per_col)) {
    i <- rand_row_per_col[[j]]
    lfc[, j] <- orig_lfc[i, j]
    pval[, j] <- orig_pval[i, j]
    stderr[, j] <- orig_stderr[i, j]
  }

  result <- SummarizedExperiment::SummarizedExperiment(
    assays = list("P.Value" = pval, "logFC" = lfc, "SE" = stderr)
  )

  return(result)
}
```

Wrapper for computing meta-results and summarizing

```{r}
# Perform and summarize the meta-results
summarizeMetaResults <- function(x) {
  meta_result <- meta_de(x, FUN = metapod::parallelFisher, pval = "P.Value", se = "SE")
  meta_result <- meta_result[!is.na(Meta.Pval)]
  meta_result[, Meta.FDR := p.adjust(Meta.Pval, method = "BH")]

  genes <- summarize_dge(
    meta_result[!Feature %like% ".*\\..*\\..*"],
    fdr = 0.1,
    lfc = log2(1.5),
    fdr_col = "Meta.FDR",
    lfc_col = "Meta.logFC"
  )

  tes <- summarize_dge(
    meta_result[Feature %like% ".*\\..*\\..*"],
    fdr = 0.1,
    lfc = log2(1.5),
    fdr_col = "Meta.FDR",
    lfc_col = "Meta.logFC"
  )

  both <- summarize_dge(
    meta_result,
    fdr = 0.1,
    lfc = log2(1.5),
    fdr_col = "Meta.FDR",
    lfc_col = "Meta.logFC"
  )

  result <- rbindlist(list("Genes" = genes, "TEs" = tes, "Both" = both), idcol = "Feature")

  return(result)
}
```

Perform the permutation meta-analysis

```{r}
run <- FALSE
if (run) {
  B <- 500
  trts2 <- c("Decitabine", "Doxorubicin", "Etoposide", "JQ1")

  meta_perms <- lapply(trts2, function(x) {
    se <- db[, db$treatment == x]
    results <- rbindlist(
      parallel::mclapply(
        1:B,
        function(i) {
          shuffled <- shuffleSE(se)
          summarizeMetaResults(shuffled)
        },
        mc.cores = 48
      ),
      idcol = "B"
    )
  })
  names(meta_perms) <- trts2
  meta_perms <- rbindlist(meta_perms, idcol = "treatment")
  fwrite(meta_perms, here(datafiles, "meta-perm-summary.tsv"), sep = "\t")
} else {
  meta_perms <- fread(here(datafiles, "meta-perm-summary.tsv"))
}
```

Display the histogram of results for each

```{r}
# Plot for genes
ggplot(meta_perms[Feature == "Genes" & Direction != "Unperturbed"], aes(x = Percent)) +
  geom_histogram(fill = "cornflowerblue", alpha = 0.9) +
  facet_grid(Direction ~ treatment) +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Protein Coding Genes",
    x = "Observed Percentage Differentially Expressed",
    y = "Number of Permutations",
    caption = "500 total permutations"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 12, color = "black"),
    axis.text = element_text(color = "black", size = 12),
    axis.title = element_text(size = 14, color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14)
  )
ggsave(here(figures, "gene-meta-perm.png"), width = 12, height = 6, dpi = 600)

# And for TEs
ggplot(meta_perms[Feature == "TEs" & Direction != "Unperturbed"], aes(x = Percent)) +
  geom_histogram(fill = "cornflowerblue", alpha = 0.9) +
  facet_grid(Direction ~ treatment) +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "TE Subfamilies",
    x = "Observed Percentage Differentially Expressed",
    y = "Number of Permutations",
    caption = "500 total permutations"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 12, color = "black"),
    axis.text = element_text(color = "black", size = 12),
    axis.title = element_text(size = 14, color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14)
  )
ggsave(here(figures, "te-meta-perm.png"), width = 12, height = 6, dpi = 600)
```

JP wants to include these results in the table included in the manuscript as 'background' levels 
of false positives. For this it might be best to compute CIs to include in the table

```{r}
perm_summary <- meta_perms[,
  .(
    avg = mean(Percent),
    ci_low = quantile(Percent, prob = 0.025),
    ci_high = quantile(Percent, prob = 0.975)
  ),
  by = .(treatment, Feature, Direction)
]

perm_summary[,
  background := paste0(round(avg, 1), "% [", round(ci_low, 1), " - ", round(ci_high, 1), "]%")
]

perm_summary <- perm_summary[, .(treatment, Feature, Direction, background)]
gene_perm_summary <- perm_summary[Feature == "Genes"][, Feature := NULL]
te_perm_summary <- perm_summary[Feature == "TEs"][, Feature := NULL]
```

Join onto the jackknife results table

```{r}
gene_jack_stats[, treatment := gsub(" \\(.*", "", Compound)]
te_jack_stats[, treatment := gsub(" \\(.*", "", Compound)]

gene_jack_stats <- merge(
  x = gene_jack_stats,
  y = gene_perm_summary,
  by = c("treatment", "Direction"),
  all.x = TRUE
)
gene_jack_stats[, treatment := NULL]
setnames(gene_jack_stats, old = "background", new = "Background [95% CI]")

te_jack_stats <- merge(
  x = te_jack_stats,
  y = te_perm_summary,
  by = c("treatment", "Direction"),
  all.x = TRUE
)
te_jack_stats[, treatment := NULL]
setnames(te_jack_stats, old = "background", new = "Background [95% CI]")
```

### Create the table showing jackknife and background estimates

```{r}
gt(gene_jack_stats, groupname_col = "Compound", row_group_as_column = TRUE) |>
  tab_header(
    title = md("Jackknife Bias Estimates for Gene Meta-Analysis")
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  cols_align(
    align = "left",
    columns = c(Compound)
  ) |>
  cols_align(
    align = "center",
    columns = c(Direction)
  ) |>
  cols_align(
    align = "right",
    columns = c(`Background [95% CI]`)
  ) |>
  fmt_percent(
    columns = c(Observed, Corrected, Bias, SE),
    decimals = 1,
    scale_values = FALSE
  ) |>
  opt_stylize(style = 1, color = "gray", add_row_striping = FALSE)
```

Same for TEs

```{r}
gt(te_jack_stats, groupname_col = "Compound", row_group_as_column = TRUE) |>
  tab_header(
    title = md("Jackknife Bias Estimates for TE Meta-Analysis")
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  cols_align(
    align = "left",
    columns = c(Compound)
  ) |>
  cols_align(
    align = "center",
    columns = c(Direction)
  ) |>
  cols_align(
    align = "right",
    columns = c(`Background [95% CI]`)
  ) |>
  fmt_percent(
    columns = c(Observed, Corrected, Bias, SE),
    decimals = 1,
    scale_values = FALSE
  ) |>
  opt_stylize(style = 1, color = "gray", add_row_striping = FALSE)
```

### Show the meta-volcano plots

```{r}
plot_gene_volcano <- function(x, ...) {
  coriell::plot_volcano(
    meta_results[[x]][!Feature %like% ".*\\..*\\..*"],
    x = "Meta.logFC",
    y = "Meta.FDR",
    fdr = 0.1,
    lfc = log2(1.5),
    raster = TRUE,
    raster_dpi = 600,
    x_axis_limits = c(-5, 5),
    y_axis_limits = c(0, 300),
    lab_digits = 0,
    ...
  ) +
    labs(
      title = x,
      subtitle = "Protein Coding Genes",
      x = "Meta logFC",
      y = "-log10(Meta FDR)",
      caption = NULL
    )
}

plot_te_volcano <- function(x, ...) {
  coriell::plot_volcano(
    meta_results[[x]][Feature %like% ".*\\..*\\..*"],
    x = "Meta.logFC",
    y = "Meta.FDR",
    fdr = 0.1,
    lfc = log2(1.5),
    raster = TRUE,
    raster_dpi = 600,
    x_axis_limits = c(-2, 2),
    y_axis_limits = c(0, 40),
    lab_digits = 0,
    up_color = "red2",
    up_shape = 1,
    ...
  ) +
    geom_point(
      data = meta_results[[x]][
        Feature %like% "^LTR.ERV" & Meta.logFC > log2(1.5) & Meta.FDR < 0.1
      ],
      aes(x = Meta.logFC, y = -log10(Meta.FDR)),
      color = "red2",
      shape = 16
    ) +
    labs(
      title = x,
      subtitle = "TE Subfamilies",
      x = "Meta logFC",
      y = "-log10(Meta FDR)",
      caption = NULL
    )
}
```

#### Patch all together

```{r}
p1 <- plot_gene_volcano("Decitabine")
p2 <- plot_te_volcano("Decitabine")
p3 <- plot_gene_volcano("Doxorubicin")
p4 <- plot_te_volcano("Doxorubicin")
p5 <- plot_gene_volcano("Etoposide")
p6 <- plot_te_volcano("Etoposide")
p7 <- plot_gene_volcano("JQ1")
p8 <- plot_te_volcano("JQ1")

# Genes
p1 + p3 + p5 + p7 + plot_layout(ncol = 2)
ggsave(here(figures, "meta-gene-volcano.png"), width = 11, height = 9, dpi = 600)
ggsave(here(figures, "meta-gene-volcano.pdf"), width = 11, height = 9)

# TEs
p2 + p4 + p6 + p8 + plot_layout(ncol = 2)
ggsave(here(figures, "meta-te-volcano.png"), width = 11, height = 9, dpi = 600)
ggsave(here(figures, "meta-te-volcano.pdf"), width = 11, height = 9)
```

### Count the ERVs that are up-regulated in each meta-analysis

```{r}
(erv_counts <- meta_results_dt[
  Meta.FDR < 0.05 &
    Meta.logFC > log2(1.5) &
    Feature %like% "^LTR.ERV" &
    treatment %chin% c("Decitabine", "Etoposide", "Doxorubicin", "JQ1"),
  .N,
  by = treatment
][order(N, decreasing = T)])
```

### What genes and TEs are shared between the sets?

```{r}
keep_trt <- c("Decitabine", "Doxorubicin", "Etoposide", "JQ1")

up_meta <- meta_results_dt[
  Meta.FDR < 0.05 & Meta.logFC > log2(1.5) & treatment %chin% keep_trt
]
down_meta <- meta_results_dt[
  Meta.FDR < 0.05 & Meta.logFC < -log2(1.5) & treatment %chin% keep_trt
]

up_meta <- split(up_meta, by = "treatment")
down_meta <- split(down_meta, by = "treatment")

# Reorder
up_meta <- up_meta[keep_trt]
down_meta <- down_meta[keep_trt]

up_genes <- lapply(up_meta, \(x) x[!Feature %like% ".*\\..*\\..*", Feature])
down_genes <- lapply(down_meta, \(x) x[Feature %like% ".*\\..*\\..*", Feature])
up_tes <- lapply(up_meta, \(x) x[Feature %like% ".*\\..*\\..*", Feature])
down_tes <- lapply(down_meta, \(x) x[Feature %like% ".*\\..*\\..*", Feature])

names(up_genes) <- names(up_tes) <- paste(names(up_genes), "(Up)")
names(down_genes) <- names(down_tes) <- paste(names(down_genes), "(Down)")

sig_genes <- c(up_genes, down_genes)
sig_tes <- c(up_tes, down_tes)
```

Create UpSet plots of the shared features in the sets of interest

```{r}
m_genes <- make_comb_mat(sig_genes, mode = "distinct")
m_tes <- make_comb_mat(sig_tes, mode = "distinct")

# limit to pairwise overlaps
m_genes <- m_genes[comb_degree(m_genes) == 2]
m_tes <- m_tes[comb_degree(m_tes) == 2]
```

```{r}
# png(
#   here(figures, "meta-upset_genes.png"),
#   width = 9,
#   height = 6,
#   units = "in",
#   res = 600
# )
pdf(here(figures, "meta-upset_genes.pdf"), width = 9, height = 6)
u1 <- UpSet(
  m_genes,
  comb_order = order(comb_size(m_genes), decreasing = TRUE),
  set_order = names(sig_genes),
  bg_col = rep(c("#F0F0FF", "#FFF0F0"), each = 4),
  top_annotation = upset_top_annotation(
    m_genes,
    add_numbers = TRUE,
    numbers_gp = gpar(cex = 1)
  ),
  right_annotation = upset_right_annotation(
    m_genes,
    add_numbers = TRUE,
    numbers_gp = gpar(cex = 1)
  ),
  column_title = "Protein Coding Genes"
)
draw(u1, padding = unit(c(1, 1, 1, 1), "cm"))
dev.off()
```

```{r}
# png(
#   here(figures, "meta-upset_tes.png"),
#   width = 9,
#   height = 6,
#   units = "in",
#   res = 600
# )
pdf(here(figures, "meta-upset_tes.pdf"), width = 9, height = 6)
u2 <- UpSet(
  m_tes,
  comb_order = order(comb_size(m_tes), decreasing = TRUE),
  set_order = names(sig_tes),
  bg_col = rep(c("#F0F0FF", "#FFF0F0"), each = 4),
  top_annotation = upset_top_annotation(
    m_tes,
    add_numbers = TRUE,
    numbers_gp = gpar(cex = 1)
  ),
  right_annotation = upset_right_annotation(
    m_tes,
    add_numbers = TRUE,
    numbers_gp = gpar(cex = 1)
  ),
  column_title = "TE Subfamilies"
)
draw(u2, padding = unit(c(1, 1, 1, 1), "cm"))
dev.off()
```

## Perform GSEA on the meta results

### Read in HALLMARK pathways

```{r}
hallmark <- msigdbr::msigdbr(
  db_species = "HS",
  species = "human",
  collection = "H"
)
hallmark <- split(as.character(hallmark$gene_symbol), hallmark$gs_name)

# Get only the interferon signatures
ifn <- hallmark[c(
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE"
)]
```

### Run CAMERA-PR on the meta stats

```{r}
run_camera <- function(x) {
  stats <- x[, Meta.z]
  names(stats) <- x[, Feature]

  idx <- limma::ids2indices(ifn, names(stats))
  result <- setDT(
    limma::cameraPR(stats, idx, use.ranks = TRUE),
    keep.rownames = "pathway"
  )

  return(list(result = result, stats = stats, idx = idx))
}

camera_results <- lapply(meta_results, run_camera)
```

Write out results

```{r}
camera_results_dt <- rbindlist(
  lapply(camera_results, `[[`, "result"),
  idcol = "treatment"
)
fwrite(
  camera_results_dt,
  here(datafiles, "meta-camerapr-results.tsv"),
  sep = "\t"
)
```

### Run fGSEA on meta z-statistics

```{r}
run_fgsea <- function(x) {
  stats <- x[, Meta.z]
  names(stats) <- x[, Feature]

  result <- fgsea::fgsea(
    pathways = ifn,
    stats = stats,
    nPermSimple = 1e4,
    eps = 0.0,
    scoreType = "pos"
  )

  return(list(result = result, stats = stats))
}

fgsea_results <- lapply(meta_results, run_fgsea)
```

Write out results

```{r}
fgsea_results_dt <- rbindlist(
  lapply(fgsea_results, `[[`, "result"),
  idcol = "treatment"
)
fwrite(fgsea_results_dt, here(datafiles, "meta-fgsea-results.tsv"), sep = "\t")
```

### Plot enrichment results

```{r}
plot_ifna <- function(x, ...) {
  coriell::plot_enrichment(
    pathway = ifn[["HALLMARK_INTERFERON_ALPHA_RESPONSE"]],
    stats = fgsea_results[[x]]$stats,
    FDR = camera_results[[x]]$result[
      pathway == "HALLMARK_INTERFERON_ALPHA_RESPONSE",
      FDR
    ],
    ES = fgsea_results[[x]]$result[
      pathway == "HALLMARK_INTERFERON_ALPHA_RESPONSE",
      ES
    ],
    NES = fgsea_results[[x]]$result[
      pathway == "HALLMARK_INTERFERON_ALPHA_RESPONSE",
      NES
    ],
    annotate = TRUE,
    anno_size = 6,
    ...
  ) +
    labs(title = x, subtitle = "Hallmark Interferon Alpha Response")
}

plot_ifny <- function(x, ...) {
  coriell::plot_enrichment(
    pathway = ifn[["HALLMARK_INTERFERON_GAMMA_RESPONSE"]],
    stats = fgsea_results[[x]]$stats,
    FDR = camera_results[[x]]$result[
      pathway == "HALLMARK_INTERFERON_GAMMA_RESPONSE",
      FDR
    ],
    ES = fgsea_results[[x]]$result[
      pathway == "HALLMARK_INTERFERON_GAMMA_RESPONSE",
      ES
    ],
    NES = fgsea_results[[x]]$result[
      pathway == "HALLMARK_INTERFERON_GAMMA_RESPONSE",
      NES
    ],
    annotate = TRUE,
    anno_size = 6,
    ...
  ) +
    labs(title = x, subtitle = "Hallmark Interferon Gamma Response")
}
```

#### Patch all together

```{r}
p1 <- plot_ifna("Decitabine")
p2 <- plot_ifny("Decitabine")
p3 <- plot_ifna("Doxorubicin")
p4 <- plot_ifny("Doxorubicin")
p5 <- plot_ifna("Etoposide")
p6 <- plot_ifny("Etoposide")
p7 <- plot_ifna("JQ1", lineColor = "grey", anno_y_pos = 0.55)
p8 <- plot_ifny("JQ1", lineColor = "grey", anno_y_pos = 0.65)

p1 + p2 + p3 + p4 + p5 + p6 + p7 + p8 + plot_layout(ncol = 4)
ggsave(here(figures, "meta-gsea.png"), width = 22, height = 9, dpi = 600)
ggsave(here(figures, "meta-gsea.pdf"), width = 22, height = 9)
```

## Effect size analysis

What is the average corelation of effect sizes within vs between treatments?

```{r}
results <- list()

for (trt in c("Doxorubicin", "Decitabine", "Etoposide", "JQ1")) {
  # Select treatments from below
  selected <- db[, db$treatment == trt]

  # Extract effect size matrices
  gene_m <- as.matrix(assay(selected[rowData(selected)$feature_type == "Gene", ], "z"))
  te_m <- as.matrix(assay(selected[rowData(selected)$feature_type == "TE", ], "z"))

  # Get NA's from SE matrix
  gene_na <- na.omit(as.matrix(assay(selected[rowData(selected)$feature_type == "Gene", ], "SE")))
  te_na <- na.omit(as.matrix(assay(selected[rowData(selected)$feature_type == "TE", ], "SE")))

  # Use only complete cases so we're not just looking at *what* was expressed
  gene_m <- gene_m[rownames(gene_na), colnames(gene_na)]
  te_m <- te_m[rownames(te_na), colnames(te_na)]

  # Perform cross-study correlations
  gene_corr <- cor(gene_m)
  te_corr <- cor(te_m)

  results[[trt]] <- list(gene = rowMeans(gene_corr), te = rowMeans(te_corr))
}
```

## Correlate TE activation with interferon signals

JP wanted to know if I could correlate TE activate with interferon response across the entire 
database. This shouldn't be too bad. For each contrast, compute the percentage of TEs that are 
up-regulated ERVs. Then compute the NES for the HALLMARK IFN A/G signatures. Plot enrichment score
as a function of percentage up-regulation.

### Compute the percentage up-regulation of TEs for all contrasts

- Should I do this for all TEs? or just ERVs? All makes sense first

```{r}
is_te <- rowData(db)$feature_type == "TE"
is_up_te <- (assay(db[is_te, ], "logFC") > 0) & (assay(db[is_te, ], "adj.P.Val") < 0.1)

# Preserve NAs so percentages are correct
is_na_te <- assay(db[is_te, ], "SE")
is_na_te[!is.na(is_na_te)] <- 1.0
is_up_te <- is_up_te * is_na_te

prop_up_te <- data.table(
  Contrast = colnames(is_up_te),
  Prop_up = colMeans(is_up_te, na.rm = TRUE)
)
```

### Compute the enrichment scores for IFN alpha/gamma

Use the z-score as the ranking stat

```{r}
pathways <- hallmark[c(
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_INFLAMMATORY_RESPONSE"
)]

# Extract z-scores and propagate NAs
gene_z_m <- as.matrix(assay(db[!is_te, ], "z"))
gene_z_m[is.na(as.matrix(assay(db[!is_te, ], "SE")))] <- NA
gene_names <- rownames(gene_z_m)

run <- TRUE
if (run) {
  ifn_results <- apply(
    X = gene_z_m,
    MARGIN = 2,
    FUN = function(x) {
      stats <- x
      names(stats) <- gene_names
      stats <- stats[!is.na(stats)]
      idx <- limma::ids2indices(pathways, names(stats))
      setDT(limma::cameraPR(stats, idx, use.ranks = TRUE), keep.rownames = "pathway")
    },
    simplify = FALSE
  )
  saveRDS(ifn_results, here(rdsfiles, "ifn-results.rds"))
} else {
  ifn_results <- readRDS(here(rdsfiles, "ifn-results.rds"))
}

ifn_results <- rbindlist(ifn_results, idcol = "Contrast")
```

### Dichotomize results

JP suggested categorizing TE enrichment by data-driven bins. Then count the enriched signatures 
within each bin.

```{r}
# Left join
ifn_props <- merge(
  x = ifn_results,
  y = prop_up_te,
  by = "Contrast",
  all.x = TRUE
)

# Create category labels for TE up-regulation
ifn_props[, te_cut := Hmisc::cut2(Prop_up, g = 3)]

# Create category labels for IFN enrichment
ifn_props[, is_enriched := fifelse(Direction == "Up" & FDR < 0.1, "Enriched", "Not Enriched")]
ifn_props[, is_enriched := factor(is_enriched, levels = c("Not Enriched", "Enriched"))]

# Now count the numbers in each category, excluding those that could not be estimated by fgsea
enrich_counts <- ifn_props[
  !is.na(is_enriched),
  .(count = .N),
  by = .(pathway, te_cut, is_enriched)
]

# Compute proportions
enrich_counts[, Prop := count / sum(count), by = .(pathway)]

# Reorder pathways for plotting
enrich_counts[,
  pathway := factor(
    pathway,
    levels = c(
      "HALLMARK_INTERFERON_ALPHA_RESPONSE",
      "HALLMARK_INTERFERON_GAMMA_RESPONSE",
      "HALLMARK_INFLAMMATORY_RESPONSE"
    )
  )
]

enrich_counts <- enrich_counts[order(pathway, te_cut, is_enriched)]

# Add labels for TE cuts
enrich_counts[,
  te_label := fcase(
    te_cut == "[0.00000,0.00274)" , "[0-0.2%)"   ,
    te_cut == "[0.00274,0.01354)" , "[0.2-1.3%)" ,
    te_cut == "[0.01354,0.96128]" , "[1.3-96%]"  ,
    default = NA
  )
][, te_label := factor(te_label, levels = c("[0-0.2%)", "[0.2-1.3%)", "[1.3-96%]"))]

# Add label for n values
enrich_counts[, n_label := paste0("n=", count)]
```

### Test associations

```{r}
# ifn alpha
ifn_alpha_tbl <- dcast(
  enrich_counts[pathway == "HALLMARK_INTERFERON_ALPHA_RESPONSE"],
  is_enriched ~ te_label,
  value.var = "count",
  fill = 0L
)
ifn_alpha_tbl <- as.matrix(ifn_alpha_tbl, rownames = "is_enriched")

# Chisq test for trend in proportions
ifn_alpha_test <- prop.trend.test(x = ifn_alpha_tbl[2, ], n = colSums(ifn_alpha_tbl))

# ifn gamma
ifn_gamma_tbl <- dcast(
  enrich_counts[pathway == "HALLMARK_INTERFERON_GAMMA_RESPONSE"],
  is_enriched ~ te_label,
  value.var = "count",
  fill = 0L
)
ifn_gamma_tbl <- as.matrix(ifn_gamma_tbl, rownames = "is_enriched")
ifn_gamma_test <- prop.trend.test(x = ifn_gamma_tbl[2, ], n = colSums(ifn_gamma_tbl))

# inflam
inflam_tbl <- dcast(
  enrich_counts[pathway == "HALLMARK_INFLAMMATORY_RESPONSE"],
  is_enriched ~ te_label,
  value.var = "count",
  fill = 0L
)
inflam_tbl <- as.matrix(inflam_tbl, rownames = "is_enriched")
inflam_test <- prop.trend.test(x = inflam_tbl[2, ], n = colSums(inflam_tbl))
```

Show plots of proportions

```{r}
# png(here(figures, "te-prop-trend.png"), width = 12, height = 5, units = "in", res = 300)
pdf(here(figures, "te-prop-trend.pdf"), width = 12, height = 5)
par(mfrow = c(1, 3), mar = c(5, 5, 4, 2) + 0.1)
# Interferon Alpha
p <- barplot(
  ifn_alpha_tbl[2, ] / colSums(ifn_alpha_tbl),
  xlab = NULL,
  ylab = "Proportion of Contrasts Enriched",
  main = "Interferon Alpha Response",
  col = "cornflowerblue",
  ylim = c(0, 0.4),
  cex.lab = 1.8,
  cex.axis = 1.5,
  cex.main = 2,
  cex.names = 1.5
)
lines(x = p, y = (ifn_alpha_tbl[2, ] / colSums(ifn_alpha_tbl)))
points(x = p, y = (ifn_alpha_tbl[2, ] / colSums(ifn_alpha_tbl)), pch = 16, cex = 1.5)
text(
  x = p,
  y = (ifn_alpha_tbl[2, ] / colSums(ifn_alpha_tbl)) + 0.02,
  labels = paste(ifn_alpha_tbl[2, ], "/", colSums(ifn_alpha_tbl)),
  cex = 1.5
)

# Interferon Gamma
p <- barplot(
  ifn_gamma_tbl[2, ] / colSums(ifn_gamma_tbl),
  xlab = "Percentage TE Up-Regulation",
  ylab = NULL,
  main = "Interferon Gamma Response",
  col = "cornflowerblue",
  ylim = c(0, 0.4),
  cex.lab = 1.8,
  cex.axis = 1.5,
  cex.main = 2,
  cex.names = 1.5
)
lines(x = p, y = (ifn_gamma_tbl[2, ] / colSums(ifn_gamma_tbl)))
points(x = p, y = (ifn_gamma_tbl[2, ] / colSums(ifn_gamma_tbl)), pch = 16, cex = 1.5)
text(
  x = p,
  y = (ifn_gamma_tbl[2, ] / colSums(ifn_gamma_tbl)) + 0.02,
  labels = paste(ifn_gamma_tbl[2, ], "/", colSums(ifn_gamma_tbl)),
  cex = 1.5
)

# Inflammatory response
# Interferon Gamma
p <- barplot(
  inflam_tbl[2, ] / colSums(inflam_tbl),
  xlab = NULL,
  ylab = NULL,
  main = "Inflammatory Response",
  col = "cornflowerblue",
  ylim = c(0, 0.4),
  cex.lab = 1.8,
  cex.axis = 1.5,
  cex.main = 2,
  cex.names = 1.5
)
lines(x = p, y = (inflam_tbl[2, ] / colSums(inflam_tbl)))
points(x = p, y = (inflam_tbl[2, ] / colSums(inflam_tbl)), pch = 16, cex = 1.5)
text(
  x = p,
  y = (inflam_tbl[2, ] / colSums(inflam_tbl)) + 0.02,
  labels = paste(inflam_tbl[2, ], "/", colSums(inflam_tbl)),
  cex = 1.5
)
dev.off()
```

Alternative plot showing all counts in 'Enriched' and 'Not enriched'. The previous plot is clearer
in it's intent. 

```{r}
# Labels for strip text
pathway_names <- c(
  "HALLMARK_INTERFERON_ALPHA_RESPONSE" = "Interferon Alpha Response",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE" = "Interferon Gamma Response",
  "HALLMARK_INFLAMMATORY_RESPONSE" = "Inflammatory Response"
)

ggplot(enrich_counts, aes(te_label, Prop, fill = is_enriched, label = n_label)) +
  geom_col(alpha = 0.9, color = "black", position = position_dodge()) +
  geom_text(position = position_dodge(width = 0.9), vjust = -1) +
  scale_fill_manual(values = c("Not Enriched" = "grey80", "Enriched" = "red2")) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 0.35)) +
  facet_wrap(~pathway, labeller = as_labeller(pathway_names)) +
  labs(x = "Percentage of TEs Up-regulated", y = "Proportion of All Contrasts", fill = NULL) +
  guides(fill = guide_legend(position = "inside")) +
  theme_classic() +
  theme(
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 18),
    strip.background = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linetype = 2),
    panel.grid.minor.y = element_line(color = "grey90", linetype = 2),
    legend.text = element_text(size = 14),
    legend.position.inside = c(0.5, 0.90)
  )
ggsave(here(figures, "te-enrichment-props.png"), width = 12, height = 6, dpi = 600)
ggsave(here(figures, "te-enrichment-props.pdf"), width = 12, height = 6)
```
