---
title: Meta-analysis of CDKis
author: Gennaro Calendo
format: html
---

Reviewers were curious about specific CDKis (rightly so). Perform meta-analysis of separate CDKis. 
Connecting to the database and filtering contrasts is the same as in the previous analysis. 

## Set up

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(coriell))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(patchwork))
suppressPackageStartupMessages(library(ComplexHeatmap))

datafiles <- here("analysis", "results", "data-files", "06")
figures <- here("analysis", "results", "figures", "06")
rdsfiles <- here("analysis", "results", "rds-files", "06")

invisible(sapply(c(datafiles, figures, rdsfiles), dir.create, showWarnings = F))
```

## Read in the dysregulation frequency data

Show same data from main figure as the first image in a multipanel image exploring CDKis in depth.

```{r}
freq_results_dt <- fread("analysis/results/data-files/04/dysregulation-frequencies.tsv")
```

Create a diverging barplot showing the TE on the y-axis and the number of contrasts where the TE is 
DE on the x-axis, splitting the up-regulated counts (in red) and down-regulated counts (in blue) in 
the middle. 

```{r}
cdki_te_freqs <- freq_results_dt[epigenetic_class_fine == "CDKi" & feature %like% "\\."][order(
  n_de,
  decreasing = TRUE
)]

cdki_te_freqs[, n_down := -1 * n_down]

cdki_te_freqs.m <- melt(
  cdki_te_freqs,
  id.vars = "feature",
  measure.vars = c("n_up", "n_down"),
  variable.name = "direction",
  value.name = "count"
)
cdki_te_freqs.m[, total := sum(abs(count)), by = feature]
setorder(cdki_te_freqs.m, -total)
cdki_te_freqs.m[, feature := factor(feature, levels = rev(unique(feature)))]
```

```{r}
ggplot(cdki_te_freqs.m[1:50], aes(x = count, y = feature, color = direction)) +
  geom_segment(
    aes(x = 0, xend = count, y = feature, yend = feature),
    linewidth = 0.5
  ) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c("n_up" = "#E41A1C", "n_down" = "#377EB8"),
    labels = c("n_up" = "Up-regulated", "n_down" = "Down-regulated")
  ) +
  scale_x_continuous(labels = abs, n.breaks = 10, limits = c(-15, 20)) +
  geom_vline(linetype = 2, xintercept = 0, color = "black", linewidth = 1) +
  theme_classic(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom",
    axis.title = element_text(size = 14)
  ) +
  labs(
    x = "Number of CDKi Contrasts Where Feature is Differentially Expressed",
    y = "Transposable Element Subfamily",
    color = "Direction of Differential Expression"
  )
ggsave(here(figures, "cdki-significant-te-contrast-freqs.pdf"), width = 9, height = 6, bg = "white")
```

## Connect to TEDEdb and filter contrasts for CDKis

Keep the filtering consistent with the previous analysis.

```{r}
db <- HDF5Array::loadHDF5SummarizedExperiment(
  "/mnt/data/data_gc/TEDEdb/appdata/se_hdf5"
)

# Remove samples from non-cancer studies
db <- db[, db$oncotree_primary_disease != "Non-Cancerous" & !is.na(db$oncotree_primary_disease)]

# Remove any samples from primary or multiple cell lines
db <- db[, !db$cell_line %like% "^PRIMARY|MULTIPLE"]

# Remove non-compound treatments
drop_mech <- c(
  "N/A",
  "small interfering rna",
  "short hairpin rna",
  "hyperthermia",
  "cas9 editing",
  "control vehicle",
  "lncrna targeting astin1",
  "lncrna targeting nkx2",
  "protac (nonfunctional)",
  "radiation therapy",
  "radiation therapy | topoisomerase inhibitor",
  "sparcle expression",
  "sparcle expression regulator",
  "sparcle expression regulator"
)
db <- db[, db$clinical_phase != "N/A" & (!db$mechanism_of_action %chin% drop_mech)]

# # Drop any contrasts that have been flagged as having outlier samples
db <- db[, db$outlier_flags == "None"]

# Also remove 'Compound Kushen Injection' since these are technically combintaions
db <- db[, !db$treatment %like% 'Compound Kushen Injection']
```

## Collect the CDKi contrasts

```{r}
cdkis <- subset(colData(db), mechanism_of_action == "cdk inhibitor")
cdkis <- setDT(data.frame(cdkis))
cdkis_by_trt <- split(cdkis, by = "treatment")

# How many contrasts in each?
cdki_counts <- sort(sapply(cdkis_by_trt, nrow), decreasing = TRUE)
trts <- names(cdki_counts)
```

What are the stats associated with these treatments?

```{r}
for (trt in trts) {
  df <- colData(db)[db$treatment == trt, ]
  msg <- paste0(
    trt,
    " - ",
    "Contrasts: ",
    nrow(df),
    "; ",
    "BioProjects: ",
    length(unique(df$experiment)),
    "; ",
    "Cell Lines: ",
    length(unique(df$cell_line)),
    "; ",
    "Lineages: ",
    length(unique(df$oncotree_lineage)),
    "; ",
    "Targets: ",
    unique(df$targets),
    "\n"
  )
  cat(msg)
}
```

## Perform meta-analysis on these data

Some have only one contrast so these are just the input data.

```{r}
meta_results_file <- here(rdsfiles, "cdki-meta-results.rds")
if (!(file.exists(meta_results_file))) {
  meta_results <- parallel::mclapply(
    trts,
    function(x) {
      se <- db[, db$treatment == x]
      result <- meta_de(
        se,
        FUN = metapod::parallelFisher,
        pval = "P.Value",
        se = "SE"
      )
      result <- result[!is.na(Meta.Pval)]
      result[, Meta.FDR := p.adjust(Meta.Pval, method = "BH")]
      return(result)
    },
    mc.cores = 16
  )
  names(meta_results) <- trts

  saveRDS(meta_results, meta_results_file)
} else {
  meta_results <- readRDS(meta_results_file)
}
```

Perform Jackknife resampling to assess stability

```{r}
process_jackknife <- function(treatment, prefix, db, run_flag) {
  message(sprintf("Running %s jacknife meta-analysis...", treatment))

  metafun <- function(x) {
    coriell::meta_de(x, metapod::parallelFisher, pval = "P.Value", se = "SE")
  }

  res <- jackknifeSE(db[, db$treatment == treatment], FUN = metafun)
  res <- rbindlist(res, idcol = "Jackknife")

  res <- res[!is.na(Meta.Pval)]
  res[, Meta.FDR := p.adjust(Meta.Pval, method = "BH"), by = Jackknife]

  return(res)
}

jk_results_file <- here(rdsfiles, "cdki-jackknife-stats.rds")
if (!(file.exists(jk_results_file))) {
  jk_results <- lapply(trts, function(x) {
    process_jackknife(
      treatment = x,
      prefix = x,
      db = db,
      run_flag = run
    )
  })
  names(jk_results) <- trts
  saveRDS(jk_results, jk_results_file)
} else {
  jk_results <- readRDS(jk_results_file)
}
```

Compute bias around the percentage estimates

```{r}
computeJackknifeBias <- function(orig, jacks, n, feature = "genes") {
  if (feature == "genes") {
    dt1 <- orig[!Feature %like% ".*\\..*\\..*"]
    dt2 <- jacks[!Feature %like% ".*\\..*\\..*"]
  } else {
    dt1 <- orig[Feature %like% ".*\\..*\\..*"]
    dt2 <- jacks[Feature %like% ".*\\..*\\..*"]
  }

  # Get the observed data from the full test
  summary_stats <- summarize_dge(
    dt1,
    fdr = 0.1,
    lfc = log2(1.5),
    fdr_col = "Meta.FDR",
    lfc_col = "Meta.logFC"
  )
  setDT(summary_stats)
  theta_hat_up <- summary_stats[Direction == "Up", Percent]
  theta_hat_down <- summary_stats[Direction == "Down", Percent]

  # Compute mean percentages across all jackknifes
  jack_stats <- dt2[,
    summarize_dge(.SD, fdr = 0.1, lfc = log2(1.5), fdr_col = "Meta.FDR", lfc_col = "Meta.logFC"),
    by = Jackknife
  ]

  jack_estimates_up <- jack_stats[Direction == "Up", Percent]
  jack_estimates_down <- jack_stats[Direction == "Down", Percent]
  theta_bar_up <- mean(jack_estimates_up)
  theta_bar_down <- mean(jack_estimates_down)

  # Compute jackknife bias and variance
  bias_jack_up <- (n - 1) * (theta_bar_up - theta_hat_up)
  bias_jack_down <- (n - 1) * (theta_bar_down - theta_hat_down)

  var_term_up <- sum((jack_estimates_up - theta_bar_up)^2)
  var_term_down <- sum((jack_estimates_down - theta_bar_down)^2)

  se_jack_up <- sqrt(((n - 1) / n) * var_term_up)
  se_jack_down <- sqrt(((n - 1) / n) * var_term_down)

  data.table(
    Direction = c("Up", "Down"),
    Observed = c(theta_hat_up, theta_hat_down),
    Corrected = c(theta_hat_up - bias_jack_up, theta_hat_down - bias_jack_down),
    Bias = c(bias_jack_up, bias_jack_down),
    SE = c(se_jack_up, se_jack_down)
  )
}
```

Collect gene and TE percentage bias esstimates for each compund and format for plotting 

```{r}
# Compute the bias and variance estimates for genes and TEs
gene_jack_stats <- Map(
  computeJackknifeBias,
  orig = meta_results[names(cdki_counts)],
  jacks = jk_results[names(cdki_counts)],
  n = cdki_counts
)
names(gene_jack_stats) <- paste0(names(gene_jack_stats), " (n=", cdki_counts, ")")
gene_jack_stats <- rbindlist(gene_jack_stats, idcol = "Compound")

te_jack_stats <- Map(
  computeJackknifeBias,
  orig = meta_results[names(cdki_counts)],
  jacks = jk_results[names(cdki_counts)],
  n = cdki_counts,
  feature = "TEs"
)
names(te_jack_stats) <- paste0(names(te_jack_stats), " (n=", cdki_counts, ")")
te_jack_stats <- rbindlist(te_jack_stats, idcol = "Compound")

# Combine both into a single dt for plotting
jack_stats <- rbindlist(
  list("Protein Coding Genes" = gene_jack_stats, "TE Subfamilies" = te_jack_stats),
  idcol = "FeatureType"
)
jack_stats[Direction == "Down", Observed := -1 * Observed]
jack_stats[is.na(Bias), SE := NaN]

# Determine the order on the plot
lvls <- jack_stats[
  FeatureType == "Protein Coding Genes" &
    Direction == "Down"
][order(Observed, decreasing = TRUE)][,
  Compound
]
jack_stats[, Compound := factor(Compound, levels = lvls)]
```

Plot the results of the meta-analysis with the SE from the Jackknife estimates

```{r}
ggplot(jack_stats, aes(x = Compound, y = Observed, fill = Direction)) +
  geom_col(color = "white") +
  geom_errorbar(data = jack_stats, aes(ymin = Observed - SE, ymax = Observed + SE), width = 0.2) +
  facet_wrap(~FeatureType) +
  scale_fill_manual(
    values = c("Up" = "#E41A1C", "Down" = "#377EB8"),
    labels = c("Up" = "Meta Up-regulated", "Down" = "Meta Down-regulated")
  ) +
  scale_y_continuous(labels = \(x) paste0(abs(x), "%"), n.breaks = 10, limits = c(-100, 100)) +
  #geom_hline(linetype = 1, yintercept = 0, color = "black", linewidth = 0.5) +
  coord_flip() +
  labs(
    fill = NULL,
    y = "Percentage of Meta-Dysregulated Features \u00B1 SE",
    x = "Compound (Number of Contrasts)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    text = element_text(color = "black"),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(linetype = 3, color = "grey80"),
    strip.text = element_text(size = 14, face = "bold")
  )
ggsave(here(figures, "cdki-meta-results-barplot.pdf"), width = 12, height = 6, bg = "white")
```

## Show meta-lfc values as a heatmap across conditions

Extract meta results as a matrix

```{r}
meta_results_dt <- rbindlist(meta_results, idcol = "Compound")

# Cast into a matrix for ComplexHeatmap
lfc_mat <- meta_results_dt[Feature %like% ".*\\..*\\..*"]
lfc_mat <- dcast(lfc_mat, Feature ~ Compound, value.var = "Meta.logFC")
lfc_mat <- as.matrix(lfc_mat, rownames = "Feature")

# Limit to TEs observed across all compounds
lfc_mat <- lfc_mat[!matrixStats::rowAnyNAs(lfc_mat), ]

# Name the columns with the number of contrassts
colnames(lfc_mat) <- paste0(colnames(lfc_mat), " (n=", cdki_counts[colnames(lfc_mat)], ")")
```

Generate a heatmap at the meta-compund level showing the TE dysregulation

```{r}
# Create the row annotation
row_anno <- tstrsplit(rownames(lfc_mat), "\\.")
te_class <- factor(row_anno[[1]], levels = c("SINE", "LINE", "LTR", "DNA", "Retroposon"))

# Create annotations for observed
cls_pal <- as.character(paletteer::paletteer_d("PNWColors::Bay"))
names(cls_pal) <- levels(te_class)

pdf(here(figures, "meta-te-lfc-heatmap.pdf"), width = 8, height = 10)
ht <- Heatmap(
  lfc_mat,
  name = "Meta logFC",
  col = circlize::colorRamp2(c(-3, 0, 3), c("#377EB8", "white", "#E41A1C")),
  right_annotation = rowAnnotation(Class = te_class, col = list(Class = cls_pal)),
  row_split = te_class,
  cluster_rows = TRUE,
  cluster_row_slices = FALSE,
  cluster_columns = TRUE,
  show_row_names = FALSE,
  row_names_gp = gpar(fontsize = 4),
  column_names_gp = gpar(fontsize = 12),
  row_title_rot = 0,
  row_gap = unit(2, "mm"),
  border = FALSE,
  column_names_rot = 45,
  column_title_gp = gpar(fontsize = 14, fontface = "bold")
)
draw(ht, padding = unit(c(20, 2, 2, 2), "mm"))
dev.off()
```

## Explore overlapping targets for each compound

```{r}
targets <- lapply(trts, function(x) {
  df <- colData(db)[db$treatment == x, ]
  unique(df$targets)
})
names(targets) <- trts

targets <- lapply(targets, \(x) strsplit(x, " | ", fixed = TRUE)[[1]])

all_targets <- sort(unique(unlist(targets)))

binary_mat <- t(sapply(targets, function(x) {
  as.integer(all_targets %in% x)
}))
colnames(binary_mat) <- all_targets

# Order the cols by target name
col_o <- gtools::mixedsort(colnames(binary_mat))

# Keep the same order as in the heatmap
ht_cols <- colnames(lfc_mat)[column_order(ht)]
ht_cols <- gsub(" \\(.*", "", ht_cols)

pdf(here(figures, "compound-target-heatmap.pdf"), width = 9, height = 6)
ht_targets <- Heatmap(
  binary_mat[ht_cols, col_o],
  name = "Target",
  col = structure(c("white", "#1F78B4"), names = c("0", "1")),
  rect_gp = gpar(col = "grey80", lwd = 1),
  cluster_rows = FALSE,
  show_row_dend = FALSE,
  cluster_columns = FALSE,
  row_names_side = "left",
  column_names_rot = 45,
  column_title = "Compound CDK Targets from Broad Repurposing Hub",
  show_heatmap_legend = FALSE
)
draw(ht_targets, padding = unit(c(2, 10, 2, 2), "mm"))
dev.off()
```

## Would it be useful to show all 63 contrasts data?

```{r}
# Extract the raw DE data from the database for all contrasts
raw_m <- assay(db[, db$treatment %chin% trts], "z")
raw_m <- raw_m[rownames(raw_m) %like% ".*\\..*\\..*", ]
raw_m <- as.matrix(raw_m)
raw_m <- raw_m[rowVars(raw_m) > 0, ]

# Remove BioProject IDs from colnames
colnames(raw_m) <- gsub("PRJNA[0-9]+\\.", "", colnames(raw_m))

# Create the row annotation
row_anno <- tstrsplit(rownames(raw_m), "\\.")
te_class <- factor(row_anno[[1]], levels = c("SINE", "LINE", "LTR", "DNA", "Retroposon"))

# Split the columns by treatment
col_trts <- db[, db$treatment %chin% trts]$treatment

# Create annotations for observed
cls_pal <- as.character(paletteer::paletteer_d("PNWColors::Bay"))
names(cls_pal) <- levels(te_class)

pdf(here(figures, "contrast-z-heatmap.pdf"), width = 16, height = 10)
ht_contrasts <- Heatmap(
  raw_m,
  name = "z-score",
  col = circlize::colorRamp2(c(-3, 0, 3), c("#377EB8", "white", "#E41A1C")),
  right_annotation = rowAnnotation(Class = te_class, col = list(Class = cls_pal)),
  row_split = te_class,
  column_split = col_trts,
  cluster_rows = TRUE,
  cluster_row_slices = FALSE,
  cluster_columns = TRUE,
  show_row_names = FALSE,
  row_names_gp = gpar(fontsize = 4),
  column_names_gp = gpar(fontsize = 10),
  row_title_rot = 0,
  row_gap = unit(2, "mm"),
  border = FALSE,
  column_names_rot = 45,
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  column_title_rot = 45
)
draw(ht_contrasts, padding = unit(c(20, 40, 20, 20), "mm"))
dev.off()
```