---
title: Meta-analysis of CDKis
author: Gennaro Calendo
format: html
---

Reviewers were curious about specific CDKis (rightly so). Perform meta-analysis of separate CDKis. 
Connecting to the database and filtering contrasts is the same as in the previous analysis. 

## Set up

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(coriell))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(patchwork))

datafiles <- here("analysis", "results", "data-files", "06")
figures <- here("analysis", "results", "figures", "06")
rdsfiles <- here("analysis", "results", "rds-files", "06")

invisible(sapply(c(datafiles, figures, rdsfiles), dir.create, showWarnings = F))
```

## Connect to TEDEdb

```{r}
db <- HDF5Array::loadHDF5SummarizedExperiment(
  "/mnt/data/data_gc/TEDEdb/appdata/se_hdf5"
)
```

## Filter contrasts

- Remove any contrasts from primary cells or multiple cell lines combined
- Remove any contrasts from non-cancer cell lines
- Remove any contrasts containing samples flagged as outliers

```{r}
db <- db[, !db$cell_line %like% "^PRIMARY|^MULTIPLE"]
db <- db[,
  db$oncotree_primary_disease != "Non-Cancerous" &
    !is.na(db$oncotree_primary_disease)
]
db <- db[, db$outlier_flags == "None"]
```

## Collect the CDKi contrasts

```{r}
cdkis <- subset(colData(db), mechanism_of_action %like% "cdk inhibitor")
cdkis <- setDT(data.frame(cdkis))
cdkis_by_trt <- split(cdkis, by = "treatment")
```

I would only like to keep single agent treatments with more than 2 contrasts present

```{r}
cdki_counts <- sapply(cdkis_by_trt, nrow)
cdki_counts <- cdki_counts[!names(cdki_counts) %flike% "+"]
cdki_counts <- sort(cdki_counts[cdki_counts > 2], decreasing = TRUE)

trts <- names(cdki_counts)
```

What are the stats associated with these treatments?

```{r}
for (trt in trts) {
  df <- colData(db)[db$treatment == trt, ]
  msg <- paste0(
    trt,
    " - ",
    "Contrasts: ",
    nrow(df),
    "; ",
    "BioProjects: ",
    length(unique(df$experiment)),
    "; ",
    "Cell Lines: ",
    length(unique(df$cell_line)),
    "; ",
    "Lineages: ",
    length(unique(df$oncotree_lineage)),
    "\n"
  )
  cat(msg)
}
```

## Perform meta-analysis on these data

```{r}
meta_results <- parallel::mclapply(
  trts,
  function(x) {
    se <- db[, db$treatment == x]
    result <- meta_de(
      se,
      FUN = metapod::parallelFisher,
      pval = "P.Value",
      se = "SE"
    )
    result <- result[!is.na(Meta.Pval)]
    result[, Meta.FDR := p.adjust(Meta.Pval, method = "BH")]
    return(result)
  },
  mc.cores = 8
)
names(meta_results) <- trts

# Save results to file
meta_results_dt <- rbindlist(meta_results, idcol = "treatment")
fwrite(
  meta_results_dt,
  here(datafiles, "cdki-meta-analysis-results.tsv"),
  sep = "\t"
)
```

## Plot the results of each meta-analysis


```{r}
plot_gene_volcano <- function(x, ...) {
  coriell::plot_volcano(
    meta_results[[x]][!Feature %like% ".*\\..*\\..*"],
    x = "Meta.logFC",
    y = "Meta.FDR",
    fdr = 0.1,
    lfc = log2(1.5),
    annotate_counts = TRUE,
    x_axis_limits = c(-5, 5),
    lab_digits = 0,
    up_shape = 16,
    up_size = 1,
    down_size = 1,
    raster = TRUE,
    raster_dpi = 600,
    ...
  ) +
    labs(
      title = x,
      subtitle = "Protein Coding Genes",
      x = "Meta logFC",
      y = "-log10(Meta FDR)",
      caption = NULL
    ) +
    theme(
      panel.grid = element_blank(),
      panel.background = element_rect(fill = "white", colour = NA)
    )
}

plot_te_volcano <- function(x, ...) {
  coriell::plot_volcano(
    meta_results[[x]][Feature %like% ".*\\..*\\..*"],
    x = "Meta.logFC",
    y = "Meta.FDR",
    fdr = 0.1,
    lfc = log2(1.5),
    annotate_counts = TRUE,
    x_axis_limits = c(-2, 2),
    lab_digits = 0,
    up_color = "red2",
    up_shape = 16,
    up_size = 1,
    down_size = 1,
    raster = TRUE,
    raster_dpi = 600,
    ...
  ) +
    geom_point(
      data = meta_results[[x]][
        Feature %like% "Alu" & Meta.logFC > log2(1.5) & Meta.FDR < 0.1
      ],
      aes(x = Meta.logFC, y = -log10(Meta.FDR)),
      color = "purple",
      shape = 16,
      size = 1.2
    ) +
    labs(
      title = x,
      subtitle = "TE Subfamilies",
      x = "Meta logFC",
      y = "-log10(Meta FDR)",
      caption = NULL
    ) +
    theme(
      panel.grid = element_blank(),
      panel.background = element_rect(fill = "white", colour = NA)
    )
}
```

Organize plots. It seems that the Alu dysregulation is mostly driven by the THZ compounds. This is
interesting. Could be intron retention / lack of splicing from broken elogation mechanisms. Would 
need library strtategy information and genomic alignments to disentangle this. 

```{r}
# Reorder plots
o <- c("THZ531", "THZ1", "HH1", "Alvocidib", "OTS964", "Palbociclib", "Abemaciclib", "Ribociclib")

gene_plots <- lapply(trts, plot_gene_volcano)
names(gene_plots) <- trts
gene_plots <- gene_plots[o]

te_plots <- lapply(trts, plot_te_volcano)
names(te_plots) <- trts
te_plots <- te_plots[o]

# Add metadata to subtitles - from above
subtitles <- c(
  "OTS964" = "Contrasts: 40; BioProjects: 1;\nCell Lines: 5; Lineages: 4",
  "THZ1" = "Contrasts: 16; BioProjects: 4;\nCell Lines: 7; Lineages: 3",
  "Palbociclib" = "Contrasts: 8; BioProjects: 3;\nCell Lines: 7; Lineages: 2",
  "Abemaciclib" = "Contrasts: 7; BioProjects: 2;\nCell Lines: 7; Lineages: 2",
  "HH1" = "Contrasts: 6; BioProjects: 1;\nCell Lines: 1; Lineages: 1",
  "Alvocidib" = "Contrasts: 5; BioProjects: 1;\nCell Lines: 5; Lineages: 1",
  "Ribociclib" = "Contrasts: 5; BioProjects: 1;\nCell Lines: 5; Lineages: 1",
  "THZ531" = "Contrasts: 5; BioProjects: 3;\nCell Lines: 4; Lineages: 3"
)
subtitles <- subtitles[o]

gene_plots <- Map(\(x, y) x + labs(subtitle = y), x = gene_plots, y = subtitles)
te_plots <- Map(\(x, y) x + labs(subtitle = y), x = te_plots, y = subtitles)
```

```{r}
pdf(here(figures, "meta-gene-volcano.pdf"), width = 22, height = 10, bg = "white")
wrap_plots(gene_plots, ncol = 4) +
  plot_annotation(title = "Protein Coding Genes") &
  theme(plot.title = element_text(size = 18, face = "bold"))
dev.off()

pdf(here(figures, "meta-te-volcano.pdf"), width = 22, height = 10, bg = "white")
wrap_plots(te_plots, ncol = 4) +
  plot_annotation(title = "TE Subfamilies") &
  theme(plot.title = element_text(size = 18, face = "bold"))
dev.off()
```